### 2020-10-30 paper.tex
Start on this

% vim: set spell spelllang=en tw=100 et sw=4 sts=4 :

\documentclass[runningheads]{llncs}

\usepackage{tikz}
\usetikzlibrary{shapes, positioning, decorations.text}
\usepackage{cleveref}

\usepackage{listings}
\lstset{
basicstyle=\small\ttfamily,
mathescape,
escapeinside={(*}{*)},
numbers=none,
keywordstyle=\bfseries,  %\color{red}
keywords = {language, Essence, given, letting, find, such, that,
            bool, int, matrix, enum, variant, record, set, mset,
            sequence, function, relation, partition, 
            domain, total, surjective, be,
            forAll, exists, sum, injective, in, preImage, range,
            new, type, intersect, union, from,
            minimising, maximising, of, indexed, by, and,
            defined, maxSize, maxNumParts, size, regular, language, cheese
            }
showstringspaces=false,
tabsize=1,
breaklines=true,
breakatwhitespace=false,
}

% \usepackage{showframe}

\title{Subgraph Isomorphism With Side Constraints}
\author{
    \"Ozg\"ur Akg\"un\inst{1} \and Jessica Enright\inst{2} \and Christopher
    Jefferson\inst{1} \and Ciaran McCreesh\inst{2} \and Patrick
    Prosser\inst{2} \and Steffen Zschaler\inst{3} \\
}
\authorrunning{\"O. Akgun et al.}
\institute{
    University of St Andrews, Scotland \and
    University of Glasgow, Scotland \\ \email{ciaran.mccreesh@glasgow.ac.uk} \and
    King's College London, England
}

\crefname{Figure}{Figure}{Figures}
\crefname{figure}{figure}{figures}

\begin{document}

\maketitle

\begin{abstract}
    The subgraph isomorphism problem is to find a small ``pattern'' graph inside a larger ``target''
    graph. There are excellent dedicated solvers for this problem, but they require substantial
    programming effort to handle complex side constraints; however, general purpose constraint
    solvers struggle on more difficult graph instances. We show how to combine the state of the art
    Glasgow Subgraph Solver with the Minion constraint programming solver to get a ``subgraphs modulo
    theories'' solver that is both performant and flexible. We also show how such an approach can be
    driven by the Essence high level modelling language. We give practical examples taken from
    ??applications.
\end{abstract}

\section{Introduction}

Finding or counting small ``pattern'' graphs inside larger ``target'' graphs is a widely applicable
hard problem, with applications including ??. This has led to the development of numerous dedicated
algorithms, with the Glasgow Subgraph Solver \cite{DBLP:conf/gg/McCreeshP020} being the current
state of the art \cite{DBLP:conf/gbrpr/Solnon19}. However, practitioners are often interested in
versions of the problem with additional restrictions, or side constraints. Some of these, such as
exact vertex labelling schemes, are trivial to include in a dedicated solver, but others currently
require either extensive programming or inefficient post-processing. This paper explores a different
approach: by allowing the Glasgow Subgraph Solver to use the Minion constraint programming (CP) solver
\cite{DBLP:conf/ecai/GentJM06} for side constraints, we achieve both the performance only a
dedicated solver can offer, with the flexibility of a full CP toolkit. This
hybrid modelling system can be driven by the Essence high level modelling language
\cite{DBLP:journals/constraints/FrischHJHM08}, making it accessible to non-specialists.

\subsection{Preliminaries}

We begin with a look at the subgraph isomorphism problem, from a high level constraint modelling
perspective.  The basic non-induced subgraph isomorphism problem is to find an injective mapping
from a pattern graph to a target graph, such that adjacent vertices in the pattern are mapped to
adjacent vertices in the target. Variations on the problem are common, and are often combined. For
example, in the induced version of the problem, non-edges must be mapped to non-edges; in the
directed version, the input graphs have directed edges whose orientations must be preserved by the
mapping; in the vertex labelled version, each vertex has a label, and the mapping must map vertices
to like-labelled vertices; and in the edge-labelled version, edges have labels which must be
preserved.  It is also common to want to count or enumerate all solutions, rather than deciding
whether at least one solution exists.  Subsets of these variations are all supported by many
dedicated subgraph isomorphism algorithms, including the Glasgow Subgraph Solver.

\begin{figure}[tb]
    \centering\begin{tikzpicture}
        \node [draw, fill] (P1) {}; \node [above=0cm of P1, font=\scriptsize] { 1 };
        \node [draw, fill, right=1cm of P1] (P2) {}; \node [above=0cm of P2, font=\scriptsize] { 2 };
        \node [draw, fill, circle, right=1cm of P2, inner sep=2.5pt] (P3) {}; \node [above=0cm of P3, font=\scriptsize] { 3 };
        \draw [->, shorten >=1mm, shorten <=1mm] (P1) -- (P2) node [midway, font=\scriptsize\itshape, above] { 1 };
        \draw [->, shorten >=1mm, shorten <=1mm] (P2) -- (P3) node [midway, font=\scriptsize\itshape, above] { 2 };
        \node [left=0.2cm of P1, anchor=east] { pattern: };

        \node [draw, fill, right=4.65cm of P1] (PT1) {};
        \node [draw, fill, circle, inner sep=2.5pt, right=0.6cm of PT1] (PT2) {};
        \draw [<->, shorten >=1mm, shorten <=1mm] (PT1) -- (PT2);
        \draw [->, shorten >=1mm, shorten <=1mm] (PT1) to [out=60, in=120, looseness=8] (PT1);

        \node [draw, fill, below=1cm of P1] (T1) {}; \node [above=0cm of T1, font=\scriptsize] { 1 };
        \node [draw, fill, right=1cm of T1] (T3) {}; \node [above=0cm of T3, font=\scriptsize] { 3 };
        \node [draw, fill, below=0.7cm of T3] (T2) {}; \node [below=0cm of T2, font=\scriptsize] { 2 };
        \node [draw, fill, circle, right=1cm of T3, inner sep=2.5pt] (T4) {}; \node [above=0cm of T4, font=\scriptsize] { 4 };
        \node [draw, fill, diamond, below=0.7cm of T4, inner sep=2pt] (T5) {}; \node [below=0cm of T5, font=\scriptsize] { 5 };
        \draw [->, shorten >=1mm, shorten <=1mm, transform canvas={yshift=0.5mm}] (T1) -- (T3) node [midway, font=\scriptsize\itshape, above] { 2 };
        \draw [<-, shorten >=1mm, shorten <=1mm, transform canvas={yshift=-0.5mm}] (T1) -- (T3) node [midway, font=\scriptsize\itshape, below] { 2 };
        \draw [->, shorten >=1mm, shorten <=1mm] (T2) -- (T3) node [midway, font=\scriptsize\itshape, left] { 1 };
        \draw [->, shorten >=1mm, shorten <=1mm] (T3) -- (T4) node [midway, font=\scriptsize\itshape, above] { 2 };
        \draw [->, shorten >=1mm, shorten <=1mm] (T3) -- (T5) node [midway, font=\scriptsize\itshape, below] { 3 };
        \node [left=0.2cm of T1, anchor=east] { target: };

        \node [draw, fill, right=5cm of T1] (TT1) {};
        \node [draw, fill, circle, inner sep=2.5pt, right=0.6cm of TT1] (TT2) {};
        \node [draw, fill, diamond, inner sep=2pt, left=0.6cm of TT1] (TT3) {};
        \draw [->, shorten >=1mm, shorten <=1mm] (TT1) -- (TT2);
        \draw [<->, shorten >=1mm, shorten <=1mm] (TT1) -- (TT3);
        \draw [->, shorten >=1mm, shorten <=1mm] (TT1) to [out=60, in=120, looseness=8] (TT1);
    \end{tikzpicture}
    \caption{A small pattern and a larger target graph used in examples throughout this paper. The
    plain text numbers are vertex names, and the shapes on vertices represent vertex labels. The
    graphs to the right are \emph{type graphs}, which are used in \cref{section:typegraphs}. The
    italic labels on edges are used for \emph{temporal graphs}, which are discussed in
    \cref{section:temporalgraphs}, and should otherwise be ignored.}
    \label{figure:littlegraphs}
\end{figure}

We can express these problems in the Essence high level modelling language, as follows. We assume
vertices take their labels from the set $L = \{ 1\ldots\ell \}$ for some given $\ell$, and edges
from $E = \{ 1\ldots{}e \}$ (and so $\ell$ and / or $e$ may be 1, for applications that do not use
labels on vertices and / or edges):
\begin{lstlisting}
given l, e : int
letting L be domain int(1..l)
letting E be domain int(1..e)
\end{lstlisting}
We take as input a pattern directed graph which has $p$ vertices (which we number from 1 to $p$, in
the set $P$), and a target directed graph which has $t$ vertices (numbered from 1 to $t$, the set
$T$). Each graph is represented as total function from vertices to vertex labels, and a
\emph{partial} function from pairs of (not necessarily distinct) vertices to edge labels:
\begin{lstlisting}
given p, t : int
letting P be domain int(1..p)
letting T be domain int(1..t)

given pat : function (P, P) --> E
given tgt : function (T, T) --> E
given plab : function (total) P --> L
given tlab : function (total) T --> L
\end{lstlisting}
Now we wish to find an injective mapping $f$:
\begin{lstlisting}
find f : function (total, injective) P --> T
\end{lstlisting}
that preserves vertex labels,
\begin{lstlisting}
such that forAll a : P .  plab(a) = tlab(f(a))
\end{lstlisting}
and directed edges, including their labels:
\begin{lstlisting}
such that forAll (a, b) in defined(pat) .
    pat((a, b)) = tgt((f(a), f(b)))
\end{lstlisting}

As a simple example, the following inputs show the problem instance represented in
\cref{figure:littlegraphs}. We have three different vertex labels (circle, square, and diamond), and
only a single edge type (which is directed; the numerical labels on edges are not used in this
section):
\begin{lstlisting}
letting l be 3
letting e be 1
\end{lstlisting}
We may now describe the pattern:
\begin{lstlisting}
letting p be 3
letting pat be function ((1, 2) --> 1, (2, 3) --> 1)
letting plab be function (1 --> 1, 2 --> 1, 3 --> 2)
\end{lstlisting}
and the target:
\begin{lstlisting}
letting t be 5
letting tgt be function ((1, 3) --> 1, (3, 1) --> 1,
  (2, 3) --> 1, (3, 4) --> 1, (3, 5) --> 1)
letting tlab be function (1 --> 1, 2 --> 1, 3 --> 1,
  4 --> 2, 5 --> 3)
\end{lstlisting}
And the Conjure tool tells us that there are exactly two solutions to the problem, as we would
expect:
\begin{lstlisting}
(1 --> 1, 2 --> 3, 3 --> 4)
(1 --> 2, 2 --> 3, 3 --> 4)
\end{lstlisting}
But what if our application requires induced isomorphisms? Then we can easily add the constraint
\begin{lstlisting}
such that forAll (a, b) : (P, P) .
    (f(a), f(b)) in defined(tgt) -> (a, b) in defined(pat)
\end{lstlisting}
And Conjure will now find us a single solution,
\begin{lstlisting}
(1 --> 2, 2 --> 3, 3 --> 4)
\end{lstlisting}
Supporting other problem variants and constraints is similarly straightforward, even if auxiliary
variables are required. For example, if instead we want to allow relabelling on vertex labels (which
is typically not supported by dedicated solvers), we could do the following:
\begin{lstlisting}
find r : function (total, injective) L --> L
such that forAll a : P .  r(plab(a)) = tlab(f(a))
\end{lstlisting}
and we would find two additional solutions,
\begin{lstlisting}
(1 --> 1, 2 --> 3, 3 --> 5)
(1 --> 2, 2 --> 3, 3 --> 5)
\end{lstlisting}
and if we removed the injective keyword for the relabelling, we would find a fifth mapping
\begin{lstlisting}
(1 --> 2, 2 --> 3, 3 --> 1)
\end{lstlisting}

\subsection{Experiments and Motivation}

\begin{figure}[tb]
\includegraphics{gen-graph-glasgow-versus-minion-cumulative.pdf}\hfill\includegraphics*{gen-graph-glasgow-versus-minion-scatter.pdf}

    \caption{Left, the cumulative number of instances solved over time, for the non-induced decision
    problem with no side constraints. Right, comparing the high level approach with the Glasgow
    Subgraph Solver on an instance by instance basis; points on the outer axes represent
    timeouts.}\label{figure:solvers}
\end{figure}

Unfortunately, whilst elegant and flexible, the performance of this approach is rather terrible on
basic subgraph isomorphism instances.  We performed computational experiments on a cluster of
??fatanodes spec, using the ??14,621 unlabelled, undirected instances from Solnon's benchmark
suite\footnote{https://perso.liris.cnrs.fr/christine.solnon/SIP.html}. This benchmark suite was
originally designed for algorithm portfolios work \cite{DBLP:conf/lion/KotthoffMS16}, and brings
together several collections of application and randomly-generated instances with varying
difficulties and solution counts (including many unsatisfiable instances). Some of the instances
have up to ?? vertices and ?? edges in patterns and up to ?? vertices and ?? edges in
targets; these lead to rather large models, by constraint programming standards.

In \cref{figure:solvers} we plot the cumulative number of instances solved over time for
the non-induced decision problem, comparing the high level approach to the Glasgow Subgraph Solver
\cite{DBLP:conf/gg/McCreeshP020} and PathLAD \cite{DBLP:conf/lion/KotthoffMS16} (the two strongest
CP approaches), and to VF2 \cite{DBLP:journals/pami/CordellaFSV04} and RI
\cite{DBLP:journals/bmcbi/BonniciGPSF13} (simpler algorithms which perform well on easy instances).
The high level approach has very slow startup times (which is to be expected as it involves
launching a Java virtual machine), but much more worryingly, only just catches up with the worst
other solver in number of instances solved as the timeout approaches. Worse, as the scatter plot in
\cref{figure:solvers} shows, there are almost no instances where the high level approach
does better than the Glasgow Subgraph Solver. And, although not pictured, the situation for the
induced variant is even more pessimistic.

These first results motivate the remainder of this paper. We want to retain the convenience of the
high level modelling approach, and to be able to add arbitrary side constraints to suit different
applications, but we do not want to have to abandon the performance that dedicated solvers can get
on hard instances. In ??section we evaluate several ways of using a CP solver in
conjunction with the Glasgow Subgraph Solver, with a focus on low level implementation details. In
??section we then return to high level modelling, and look at the convenience it provides for
??temporal problems, ??relabelling and retyping, and ??optimisation problems.

\section{Hybrid Solving}

The Glasgow Subgraph Solver employs a CP approach to solve subgraph-finding problems, but using
special data structures and algorithms---for example, rather than representing the adjacency
constraint using a table, it uses bitset adjacency matrices. The solver also exploits various graph
invariants involving degrees and paths to further reduce the search space, and also has special
search order heuristics. From this paper's perspective, the most important design aspect is that
internally, the solver has a CP style variable for each vertex in the pattern graph, whose domains
range over the vertices of the target graph. The solver performs a backtracking search\footnote{The
solver also makes use of restarts and nogood recording, rather than just na\"\i{}ve backtracking,
but this does not hugely affect this paper.}
attempting to assign each variable a value from its domain, whilst respecting adjacency and
injectivity constraints. At each recursive call of search, the solver performs \emph{propagation} to
eliminate infeasible values from domains. If any domain becomes empty, the solver backtracks;
otherwise, it selects a variable, and tries assigning it each value from its domain in turn.

For now we will assume that there is some way of setting up the subgraph solver and a CP solver such
that they both have this same set of variables and values, and that there is some way of knowing how
to form a correspondence between their internal representations. We will allow the CP solver to have
additional variables that the subgraph solver does not know about (such as the second mapping used
in the relabelling example in the previous section), and we do not specifically require the CP
solver to be aware of all of the graph constraints.  This gives us several potential basic
approaches to handling side constraints:
\begin{itemize}
    \item We could use a CP solver as a \emph{solution checker}. Whenever the subgraph solver finds
        a solution, it will pass it to the CP solver, which will treat the solution as a set of
        equality constraints. The CP solver will then attempt to find a satisfying assignment. If
        the CP solver does not have any additional variables, this is equivalent to simply checking
        that the remaining constraints hold, but otherwise it can involve search. For a decision
        problem, the CP solver then communicates back to the subgraph solver either ``yes, this is a
        valid solution'', or ``no, reject this solution and keep going''. If we
        are solving a counting or enumeration problem, the CP solver must find \emph{all} solutions
        and communicate this back to the subgraph solver.
    \item We could additionally ask a CP solver at every stage of search to \emph{test} whether the
        state the subgraph solver is in is obviously infeasible. Whenever the subgraph solver has
        finished performing propagation, it can communicate the \emph{trail} (that is, its current
        sequence of guessed assignments) to the CP solver, which again treats these as additional
        equality constraints. The CP solver then performs its own propagation (but not search), and
        communicates back either a ``yes, keep going'' or a ``no, backtrack immediately''.
    \item Finally, after this testing, we could also ask the CP solver to communicate any deletions
        it infers back to the subgraph solver. In other words, the subgraph solver would use the CP
        solver as an additional propagator.
\end{itemize}
Unfortunately, each of these approaches has drawbacks. The solution we will settle upon is based
upon \emph{rollbacks}; we will describe this below, after presenting experiments that demonstrate
the difference between these approaches.

\subsection{Implementation}

?? How we do this: FIFOs, simple text based protocol. Both solvers are run and initialised. The
subgraph solver calls the CP solver as a propagator. The CP solver sets up the model, then clones it
and adds in new constraints for each query.

\subsection{Experiments}

\begin{figure}[p]
    \includegraphics{gen-graph-nosideconstraints.pdf}\hfill\includegraphics{gen-graph-oddeven.pdf}
    \medskip
    \includegraphics{gen-graph-mostlyodd.pdf}\hfill\includegraphics{gen-graph-parity.pdf}
    \caption{Comparing different approaches to hybrid solving. On the top row, ``no side
    constraints'' then with the ``more odd target vertices than even target vertices'' side
    constraint; on the bottom row, the ``mostly odd target vertices'' side constraint on the left,
    and the ``odd to odd, even to even'' side constraint on the right.}\label{figure:cumulative}
\end{figure}

We now present the results of some computational experiments. The experiments in this section are
designed to be hard, and to emphasise the difference between the approaches, rather than to be
realistic. We will continue to work with Solnon's non-induced subgraph isomorphism benchmark
instances, but will consider three variations. Firstly, we will consider the problem with no side
constraints. This is, in some sense, the worst case scenario, where we must pay the full price of
hybrid solving, but cannot get any benefit from it. Secondly, let's say that the number of odd
target vertices used must be greater than the number of even target vertices used:
\newcommand{\modsymbol}{\%}
\begin{lstlisting}
such that (sum i : P . f(i) (*\modsymbol{}*) 2) >
    (sum i : P . (f(i) + 1) (*\modsymbol{}*) 2)
\end{lstlisting}
Thirdly, let us say that fewer than three odd target vertices may be used:
\begin{lstlisting}
such that (sum i : P . f(i) (*\modsymbol{}*) 2) < 3
\end{lstlisting}
And fourthly, let us say that even pattern vertices must be mapped to even target vertices, and odd
pattern vertices to odd target vertices:
\begin{lstlisting}
such that forAll a : P . (a (*\modsymbol{}*) 2) = (f(a) (*\modsymbol{}*) 2)
\end{lstlisting}

We chose these last three problems because intuitively, the first is likely not to be able to
perform inference until deep in a search tree (when most pattern vertices are mapped to specific
target vertices), the second is likely to be able to perform inference early in the search tree
(after a few assignments have been made, but not at the root node), and the third will propagate
only at the root node.

The results of these experiments are presented in
\cref{figure:cumulative}. Let us first look at the top left plot, where we do not actually have any side constraints. When calling
the CP solver as a solution checker, we ultimately achieve the same performance as the Glasgow
Subgraph Solver\footnote{Actually, because high level modelling can use different names for
variables and values, we get slight differences due to changes to tiebreaking in search order
heuristics.}, although we can pay a substantial startup overhead. This should not be surprising:
with no side constraints, the subgraph solver runs as normal, and will perform just one call to the
CP solver on satisfiable instances. The testing and propagating approaches are both over an order of
magnitude slower: calling the CP solver at every search node is clearly very expensive---although if
we do call the CP solver, having it communicate deletions is not much of an additional overhead.
Finally, all approaches substantially outperform using a CP solver on its own without help from the
subgraph solver.

?? Include a version where Minion doesn't actually get the constraints

What about the remaining three plots in \cref{figure:cumulative}, where we do have side constraints (and so can no
longer compare to the subgraph solver on its own)? As we hoped, we see differences between the three
plots. On the top right, where we expect the side constraints to fire late, solution
checking clearly beats testing or propagation during search. However, on the bottom left,
where we expect side constraints to fire early, the testing and propagating approaches are much
better than solution checking. ?? Bottom right. Finally, any approach using the subgraph solver
remains much better than a pure CP approach.

\subsection{A Rollback Approach}

From what we have seen so far, it is obviously important to call the CP solver \emph{some} of the
time during search, but too expensive to call it \emph{all} of the time. We will therefore introduce
a new approach, which we call \emph{rollback}. This approach is inspired by backjumping ??cite, as
well as by the conflict analysis methods used in SAT and SMT solvers ??cite and in lazy clause
generating CP solvers ??cite. The idea is as follows.  Firstly, we call the CP solver with full
propagation at the root node, in case we are dealing with a particularly rich labelling scheme.
Secondly, we use the CP solver for solution checking, since this is required for correctness. Now,
suppose the CP solver rejects a candidate solution: this will cause the subgraph solver to
backtrack. At this point, we call the CP solver again, with full propagation. Either the CP solver
is happy, in which case we proceed with search, or the CP solver indicates failure, in which case we
backtrack again, and do another attempt at full propagation, and so on until the CP solver is happy.

The idea behind this approach is to avoid calling the CP solver when it is unlikely to do anything
useful, but that once a failure has been encountered, we want to extract as much information as we
can from the CP solver. If the failure encountered was due to a ``local'' property of the solution,
such as in the ``more odd than even'' example, then we will quickly return to just using the
subgraph solver for search. However, if the failure is due to only a few early assignments, as in
the ``fewer than three odd vertices'' example, then we will jump back to nearly the root of the
search tree.

The results in \cref{figure:nosideconstraints,figure:oddeventhreeodd} demonstrate the
success of this approach. When there are no side constraints, this approach has no overheads
compared to solution checking. When constraints fire late, this approach is slightly better than
solution checking, and when constraints fire early, this approach is slightly better than always
testing or propagating during search. In other words, rolling back from failures gives us all of the
strengths and none of the weaknesses of the simpler approaches.

?? Randomly also calling in unsat trees?

\section{Subgraph Problems with Side Constraints}

\subsection{Retyping Problems}\label{section:typegraphs}

A \emph{typed graph} is a kind of labelled graph where the labels also carry a graph structure
??cite. In general, mappings between typed graphs must preserve the label structure, but here we
will be looking at a problem variation involving relabelling which occurs in ??application. We may
describe typed graph subisomorphism problems in Essence as follows. As before, we are given a
pattern graph and a target graph, both of which carry labels; we will draw the vertex labels from
different sets, to emphasise
the relabelling.
\begin{lstlisting}
given pl, tl, e : int
letting PL be domain int(1..pl)
letting TL be domain int(1..tl)
letting E be domain int(1..e)
\end{lstlisting}
We are also given labelled graphs,
\begin{lstlisting}
given p, t : int
letting P be domain int(1..p)
letting T be domain int(1..t)

given pat : function (P, P) --> E
given tgt : function (T, T) --> E
given plab : function (total) P --> PL
given tlab : function (total) T --> TL
\end{lstlisting}
but now the labels also carry a graph structure,
\begin{lstlisting}
given pattype : function (PL, PL) --> E
given tgttype : function (TL, TL) --> E
\end{lstlisting}
We are looking for an injective mapping from the pattern graph to the target graph,
\begin{lstlisting}
find f : function
    (total, injective) P --> T
\end{lstlisting}
as well as an injective mapping between the label graphs,
\begin{lstlisting}
find r : function
    (total, injective) PL --> TL
\end{lstlisting}
in such a way that graph structure and labels are preserved,
\begin{lstlisting}
such that forAll (a,b) in defined(pat) .
    pat((a,b)) = tgt((f(a), f(b)))
such that forAll a : P .
    r(plab(a)) = tlab(f(a))
\end{lstlisting}
and also requiring that the structure on the labels is preserved,
\begin{lstlisting}
such that forAll (a,b) in defined(pattype) .
    pattype((a,b)) = tgttype((r(a),r(b)))
\end{lstlisting}

Consider again the example in \cref{figure:littlegraphs}, and now suppose they are equipped with the type
structures shown to the right of each graph,
\begin{lstlisting}
letting pattype be function (
  (1, 1) --> 1, (1, 2) --> 1, (2, 1) --> 1 )
letting tgttype be function (
  (1, 1) --> 1, (1, 2) --> 1,
  (1, 3) --> 1, (3, 1) --> 1 )
\end{lstlisting}
Recall that with vertex relabelling, four solutions exist. However, with retyping, only two of the
solutions respect the type graph,
\begin{lstlisting}
(1 --> 1, 2 --> 3, 3 --> 5)
(1 --> 2, 2 --> 3, 3 --> 5)
\end{lstlisting}
whilst the solutions that instead map pattern vertex 3 to target vertex 4 are no longer valid.

\subsection{Temporal Subgraph Problems}\label{section:temporalgraphs}

In a \emph{temporal graph}, edges are labelled with some kind of timestamp---here we will use
integers. Temporal pattern matching and counting has applications in ??. There are at least three
common kinds of temporal subgraph isomorphism. In an \emph{exact} subisomorphism, times are simply
labels that must match exactly. If we look at \cref{figure:littlegraphs}, now ignoring vertex labels
but using the edge labels,
\begin{lstlisting}
letting l be 1
letting e be 3

letting p be 3
letting pat be function ((1, 2) --> 1, (2, 3) --> 2)
letting plab be function (1 --> 1, 2 --> 1, 3 --> 1)

letting t be 5
letting tgt be function ((1, 3) --> 2, (3, 1) --> 2,
  (2, 3) --> 1, (3, 4) --> 2, (3, 5) --> 3)
letting tlab be function (1 --> 1, 2 --> 1, 3 --> 1,
  4 --> 1, 5 --> 1 )
\end{lstlisting}
then there are two solutions,
\begin{lstlisting}
(1 --> 2, 2 --> 3, 3 --> 1)
(1 --> 2, 2 --> 3, 3 --> 4)
\end{lstlisting}

A less strict kind of subisomorphism is an \emph{offset}, where edge labels must match exactly but
offset by an integer constant. In our example, this means ``find a mapping where the event from 2
to 3 occurs one time unit after the event from 1 and 2''. We can model this as follows:
\begin{lstlisting}
find o : int(-e..e)
such that forAll (a,b) in defined(pat) .
    pat((a,b)) = o + tgt((f(a), f(b)))
\end{lstlisting}
and we find one additional solution,
\begin{lstlisting}
(1 --> 1, 2 --> 3, 3 --> 5)
\end{lstlisting}

Finally, in an \emph{order} embedding, the pattern edge labels simply define an order on events. We
can model this as follows:
\begin{lstlisting}
find o : function (total) E --> E
such that forAll x : int(1..e - 1) .  o(x) <= o(x + 1)
such that forAll (a,b) in defined(pat) .
    pat((a,b)) = o(tgt((f(a), f(b))))
\end{lstlisting}
which gets us yet another solution,
\begin{lstlisting}
(1 --> 2, 2 --> 3, 3 --> 5)
\end{lstlisting}

\subsection{Compiler Instruction Generation}

\subsection{Subgraph Isomorphism With Costs}

We can also support optimisation problems. If, for example, each target vertex has a cost associated with it,
\begin{lstlisting}
given tcost : function (total) T --> int
\end{lstlisting}
then we can ask to find the cheapest solution,
\begin{lstlisting}
minimising sum([ tcost(f(a)) | a : P])
\end{lstlisting}

\section{Conclusion}

\bibliographystyle{splncs04}
\bibliography{paper}

\end{document}

### 2020-10-30 experiments/ladtoessence.cc
Experiments

@@ -1,0 +1,69 @@
+/* vim: set sw=4 sts=4 et foldmethod=syntax : */
+
+#include <fstream>
+#include <iostream>
+#include <string>
+
+using std::cout;
+using std::endl;
+using std::ifstream;
+using std::string;
+
+auto convert(const bool notable, const string & filename, const string & nodes, const string & labels, const string & lfn,
+        const string & fn) -> void
+{
+    ifstream infile{ filename };
+    int nv;
+    infile >> nv;
+    cout << "letting " << nodes << " be " << nv << endl;
+    cout << "letting " << labels << " be 1" << endl;
+
+    if (! notable) {
+        cout << "letting " << fn << " be function" << endl;
+
+        bool first = true;
+        for (int i = 0 ; i < nv ; ++i) {
+            int ne;
+            infile >> ne;
+            for (int j = 0 ; j < ne ; ++j) {
+                int t;
+                infile >> t;
+
+                if (first) {
+                    cout << "( ";
+                    first = false;
+                }
+                else
+                    cout << ", ";
+                cout << "(" << (i + 1) << "," << (t + 1) << ") --> 1" << endl;
+            }
+        }
+        cout << ")" << endl;
+    }
+
+    cout << "letting " << lfn << " be function" << endl;
+    bool first = true;
+    for (int i = 0 ; i < nv ; ++i) {
+        if (first) {
+            cout << "( ";
+            first = false;
+        }
+        else
+            cout << ", ";
+        cout << (i + 1) << " --> 1";
+    }
+    cout << ")" << endl;
+}
+
+auto main(int argc, char * argv[]) -> int
+{
+    if (3 != argc && 4 != argc)
+        return EXIT_FAILURE;
+
+    bool notable = (argc == 4 && string{ argv[1] } == "--notable");
+
+    convert(notable, argv[1 + notable], "p", "l", "plab", "pat");
+    convert(notable, argv[2 + notable], "t", "e", "tlab", "tgt");
+    return EXIT_SUCCESS;
+}
+

### 2020-10-30 experiments/run-experiment.sh

@@ -1,0 +1,90 @@
+#!/usr/bin/env bash
+
+results=$1
+alg=$2
+instance=$3
+pattern=$4
+target=$5
+
+problem=$(echo ${alg} | sed -n -e 's,.*si-\([^-]*\)-.*,\1,p' )
+solver=
+MS_SR_OPTS=
+MS_GSS_OPTS=
+if echo ${alg} | grep -q -- '-gss-' ; then
+    solver=gss
+else
+    if echo ${alg} | grep -q -- '-minion-' ; then
+        solver=minion
+    elif echo ${alg} | grep -q -- '-hybrid-' ; then
+        solver=hybrid
+        if echo ${alg} | grep -q -- '-comm-checker-' ; then
+            MS_GSS_OPTS='--propagate-using-lackey never'
+        elif echo ${alg} | grep -q -- '-comm-test-' ; then
+            MS_GSS_OPTS='--propagate-using-lackey never --send-partials-to-lackey'
+        elif echo ${alg} | grep -q -- '-comm-propagate-' ; then
+            MS_GSS_OPTS='--propagate-using-lackey always --send-partials-to-lackey'
+        elif echo ${alg} | grep -q -- '-comm-rollback-' ; then
+            MS_GSS_OPTS='--propagate-using-lackey root-and-backjump'
+        fi
+    fi
+
+    if echo ${alg} | grep -q -- '-preprocess-none-' ; then
+        MS_SR_OPTS='-preprocess None'
+    elif echo ${alg} | grep -q -- '-preprocess-gac-' ; then
+        MS_SR_OPTS='-preprocess GAC'
+    fi
+fi
+
+[[ -n $problem ]] && [[ -n $solver ]] || exit 1
+
+export MS_SR_OPTS
+export MS_GSS_OPTS
+
+mkdir -p ${results}/${alg}/${instance} || exit 1
+cd ${results}/${alg}/${instance} || exit 1
+
+cat <<END > run-experiment.params
+alg=$alg
+problem=$problem
+solver=$solver
+pattern=$pattern
+target=$target
+instance=$instance
+MS_GSS_OPTS=$MS_GSS_OPTS
+MS_SR_OPTS=$MS_SR_OPTS
+END
+
+kill_descendants ()
+{
+    local children=$(ps -o pid= --ppid "$1")
+    for pid in $children ; do
+        kill_descendants "$pid"
+    done
[…]

### 2020-10-30 code/minion-subgraph.sh
Magic subgraph runner

@@ -1,0 +1,27 @@
+#!/usr/bin/env bash
+
+if [[ "$*" == *outputCompressedDomains* ]] ; then
+    exec minion $*
+fi
+
+if [[ "$*" == *findallsols* ]] ; then
+    GRAPHFLAGS="--print-all-solutions"
+fi
+
+rm -f fifo.A fifo.B
+mkfifo fifo.A
+mkfifo fifo.B
+
+minion $MS_MINION_OPTS $* -command-list fifo.A fifo.B &
+
+glasgow_subgraph_solver $GRAPHFLAGS \
+    --pattern-format csvname:find-f.csv --target-format csv \
+    given-pat.csv given-tgt.csv \
+    --send-to-lackey fifo.A --receive-from-lackey fifo.B \
+    $MS_GSS_OPTS \
+    | tee gss.out
+ret=$?
+
+wait
+
+exit $?

### 2020-10-30 experiments/runtimes.cc
Runtimes extraction

@@ -1,0 +1,78 @@
+/* vim: set sw=4 sts=4 et foldmethod=syntax : */
+
+#include <vector>
+#include <filesystem>
+#include <string>
+#include <fstream>
+#include <sstream>
+#include <iostream>
+#include <map>
+#include <cstdlib>
+
+using std::cout;
+using std::endl;
+using std::vector;
+using std::filesystem::path;
+using std::string;
+using std::stringstream;
+using std::getline;
+using std::map;
+using std::stold;
+using std::ifstream;
+
+auto main(int argc, char * argv[]) -> int
+{
+    vector<path> dirs;
+    for (int p = 1 ; p < argc ; ++p)
+        dirs.emplace_back(argv[p]);
+
+    {
+        bool first = true;
+        for (auto & d : dirs) {
+            if (! first)
+                cout << " ";
+            cout << d.filename();
+            first = false;
+        }
+        cout << endl;
+    }
+
+    ifstream instances{ "instances.txt" };
+    string line;
+    while (getline(instances, line)) {
+        stringstream line_s{ line };
+        string n, p, t, f;
+        if (! (line_s >> n >> p >> t >> f))
+            return EXIT_FAILURE;
+
+        cout << n;
+        for (auto & d : dirs) {
+            path p = d / n / "run-experiment.out";
+            ifstream result{ p };
+            map<string, string> entries;
+            string result_line;
+            while (getline(result, result_line)) {
+                auto eq = result_line.find('=');
+                if (eq == string::npos)
+                    return EXIT_FAILURE;
+                entries.emplace(result_line.substr(0, eq), result_line.substr(eq + 1));
+            }
+
+            if (! (entries.count("started") && entries.count("finished") && entries.count("retcode")))
+                return EXIT_FAILURE;
+
+            if (entries["retcode"] == "0") {
+                long double s = stold(entries["started"]), f = stold(entries["finished"]);
+                cout << " " << (f - s);
+            }
+            else if (entries["retcode"] == "1") {
+                cout << " NaN";
+            }
+            else
+                return EXIT_FAILURE;
+        }
+        cout << endl;
+    }
+
+    return EXIT_SUCCESS;
+}

### 2020-10-31 experiments/runtimes.cc
Correct header

@@ -26,16 +26,10 @@ auto main(int argc, char * argv[]) -> int
     for (int p = 1 ; p < argc ; ++p)
         dirs.emplace_back(argv[p]);
 
-    {
-        bool first = true;
-        for (auto & d : dirs) {
-            if (! first)
-                cout << " ";
-            cout << d.filename();
-            first = false;
-        }
-        cout << endl;
-    }
+    cout << "instance";
+    for (auto & d : dirs)
+        cout << " " << d.filename();
+    cout << endl;
 
     ifstream instances{ "instances.txt" };
     string line;

### 2020-11-16 pathLAD/allDiff.c
Bring in LAD

@@ -1,0 +1,339 @@
+// This software has been written by Christine Solnon.
+// It is distributed under the CeCILL-B FREE SOFTWARE LICENSE
+// see http://www.cecill.info/licences/Licence_CeCILL-B_V1-en.html for more details
+
+#define white 0
+#define grey 1
+#define black 2
+#define toBeDeleted 3
+#define deleted 4
+
+void addToDelete(int u, int* list, int* nb, int* marked){
+	if (marked[u]<toBeDeleted){
+		list[(*nb)++]=u;
+		marked[u]=toBeDeleted; 
+	}
+}
+
+bool updateMatching(int sizeOfU, int sizeOfV, int* degree, int* firstAdj, int*  adj, int* matchedWithU){
+	// input:
+	// sizeOfU = number of vertices in U
+	// sizeOfV = number of vertices in V
+	// degree[u] = number of vertices of V which are adjacent to u
+	// firstAdj[u] = pos in adj of the first vertex of V adjacent to u
+	// adj[firstAdj[u]..firstAdj[u]+sizeOfU[u]-1] = vertices of V adjacent to u
+	
+	// input/output:
+	// matchedWithU[u] = vertex of V matched with u
+	
+	// returns true if there exists a matching that covers U, i.e., if for every u in 0..nbU-1, 
+	// there exists a different v in 0..nb-1 such that v is adjacent to u;
+	// returns false otherwise
+	
+	if (sizeOfU>sizeOfV) return false; // trivial case of infeasibility
+	
+	int matchedWithV[sizeOfV]; // matchedWithV[matchedWithU[u]]=u
+	int nbPred[sizeOfV]; // nbPred[i] = nb of predecessors of the ith vertex of V in the DAG
+	int pred[sizeOfV][sizeOfU]; // pred[i][j] = jth predecessor the ith vertex of V in the DAG
+	int nbSucc[sizeOfU]; // nbSucc[i] = nb of successors of the ith vertex of U in the DAG
+	int succ[sizeOfU][sizeOfV]; // succ[i][j] = jth successor of the ith vertex of U in the DAG
+	int listV[sizeOfV], listU[sizeOfU], listDV[sizeOfV], listDU[sizeOfU];
+	int nbV, nbU, nbDV, nbDU;
+	int i,j,k,stop,u,v;
[…]

### 2020-11-16 vflib/solve_vf_induced.cc
VF solver

@@ -1,0 +1,153 @@
+/* vim: set sw=4 sts=4 et foldmethod=syntax : */
+
+#include "argraph.h"
+#include "argedit.h"
+#include "argloader.h"
+#include "match.h"
+#include "vf2_sub_state.h"
+
+#include <fstream>
+#include <iostream>
+#include <thread>
+#include <condition_variable>
+#include <mutex>
+#include <atomic>
+#include <cstdlib>
+#include <chrono>
+#include <vector>
+
+using std::chrono::milliseconds;
+using std::chrono::steady_clock;
+using std::chrono::duration_cast;
+
+namespace
+{
+    void load(const std::string & file, ARGEdit & ed, std::vector<int> & loops)
+    {
+        std::ifstream f(file.c_str());
+
+        int nv;
+        f >> nv;
+        loops.resize(nv);
+        for (int i = 0 ; i < nv ; ++i)
+            ed.InsertNode(&loops[i]);
+
+        for (int r = 0 ; r < nv ; ++r) {
+            int nn;
+            f >> nn;
+            for (int n = 0 ; n < nn ; ++n) {
+                int nb;
+                f >> nb;
+                if (r == nb)
+                    loops[r] = true;
+                else
+                    ed.InsertEdge(r, nb, NULL);
+            }
+        }
+
+        if (! f)
+            throw 0;
+    }
+
+    struct OurComparator : AttrComparator
+    {
+        virtual bool compatible(void *attr1, void *attr2)
+        {
+            return *static_cast<int *>(attr1) == *static_cast<int *>(attr2);
+        }
+    };
+}
+
+int main(int argc, char * argv[])
+{
+    if (argc != 4)
+        return EXIT_FAILURE;
+
+    std::ifstream pattern_in(argv[1], std::ios::in | std::ios::binary);
+    std::ifstream target_in(argv[2], std::ios::in | std::ios::binary);
+
+    int timeout = atoi(argv[3]);
+
+    ARGEdit pattern_ed, target_ed;
+    std::vector<int> pattern_loops, target_loops;
+
+    load(argv[1], pattern_ed, pattern_loops);
+    load(argv[2], target_ed, target_loops);
+
+    int n;
+    node_id ni1[8000], ni2[8000];
+
+    std::thread timeout_thread;
+    std::mutex timeout_mutex;
+    std::condition_variable timeout_cv;
+
+    std::atomic<int> state;
+    state.store(0);
+
+    /* Start the clock */
[…]

### 2020-11-16 ri/lad2gfu.cc
RI runner

@@ -1,0 +1,111 @@
+/* vim: set sw=4 sts=4 et foldmethod=syntax : */
+
+#include <algorithm>
+#include <cstdlib>
+#include <exception>
+#include <iostream>
+#include <numeric>
+#include <string>
+#include <vector>
+
+using std::accumulate;
+using std::cerr;
+using std::cin;
+using std::cout;
+using std::endl;
+using std::exception;
+using std::istream;
+using std::string;
+using std::vector;
+
+class GraphFileError :
+    public exception
+{
+    private:
+        string _what;
+
+    public:
+        GraphFileError(const string & filename, const string & message) throw ();
+
+        auto what() const throw () -> const char *;
+};
+
+GraphFileError::GraphFileError(const string & filename, const string & message) throw () :
+    _what("Error reading graph file '" + filename + "': " + message)
+{
+}
+
+auto GraphFileError::what() const throw () -> const char *
+{
+    return _what.c_str();
+}
+
+namespace
+{
+    auto read_word(istream & infile) -> int
+    {
+        int x;
+        infile >> x;
+        return x;
+    }
+}
+
+auto main(int, char * argv[]) -> int
+{
+    try {
+        uint16_t size = read_word(cin);
+        if (! cin)
+            throw GraphFileError{ "stdin", "error reading size" };
+
+        vector<vector<int> > adj;
+        adj.resize(size, vector<int>(size));
+
+        for (int r = 0 ; r < size ; ++r) {
+            int c_end = read_word(cin);
+            if (! cin)
+                throw GraphFileError{ "stdin", "error reading edges count" };
+
+            for (int c = 0 ; c < c_end ; ++c) {
+                int e = read_word(cin);
+
+                if (e < 0 || e >= size)
+                    throw GraphFileError{ "stdin", "edge index out of bounds" };
+
+                adj[r][e] = adj[e][r] = 1;
+            }
+        }
+
+        string rest;
+        if (cin >> rest)
+            throw GraphFileError{ "stdin", "EOF not reached, next text is \"" + rest + "\"" };
+        if (! cin.eof())
+            throw GraphFileError{ "stdin", "EOF not reached" };
+
+        cout << "#graph" << endl;
+        cout << size << endl;
+        for (int r = 0 ; r < size ; ++r)
+            cout << (adj[r][r] ? "loop" : "noloop") << endl;
+
+        int ec = 0;
[…]

### 2020-11-16 ri/solve_ri3.cc

@@ -1,0 +1,302 @@
+/*
+Copyright (c) 2014 by Rosalba Giugno
+
+This library contains portions of other open source products covered by separate
+licenses. Please see the corresponding source files for specific terms.
+
+RI is provided under the terms of The MIT License (MIT):
+
+Permission is hereby granted, free of charge, to any person obtaining a copy of
+this software and associated documentation files (the "Software"), to deal in
+the Software without restriction, including without limitation the rights to
+use, copy, modify, merge, publish, distribute, sublicense, and/or sell copies of
+the Software, and to permit persons to whom the Software is furnished to do so,
+subject to the following conditions:
+
+The above copyright notice and this permission notice shall be included in all
+copies or substantial portions of the Software.
+
+THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
+AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY,
+WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
+CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE SOFTWARE.
+ */
+
+#include <iostream>
+#include <fstream>
+#include <string>
+#include <cstdlib>
+#include <ctime>
+
+
+#include <stdio.h>
+#include <stdlib.h>
+#include "c_textdb_driver.h"
+#include "timer.h"
+
+
+#include "AttributeComparator.h"
+#include "AttributeDeconstructor.h"
+#include "Graph.h"
+#include "MatchingMachine.h"
+#include "MaMaConstrFirst.h"
+#include "Match.h"
+#include "Solver.h"
+#include "IsoGISolver.h"
+#include "SubGISolver.h"
+#include "InducedSubGISolver.h"
+
+#define PRINT_MATCHES
+//#define CSV_FORMAT
+
+
+using namespace rilib;
+
+struct Success
+{
+};
+
+class FirstOnlyConsoleMatchListener : public MatchListener {
+public:
[…]

### 2020-11-16 experiments/run-experiment.sh
More experiment runners

@@ -12,6 +12,12 @@ MS_SR_OPTS=
 MS_GSS_OPTS=
 if echo ${alg} | grep -q -- '-gss-' ; then
     solver=gss
+elif echo ${alg} | grep -q -- '-pathlad'- ; then
+    solver=pathlad
+elif echo ${alg} | grep -q -- '-vf2'- ; then
+    solver=vf2
+elif echo ${alg} | grep -q -- '-ri'- ; then
+    solver=ri
 else
     if echo ${alg} | grep -q -- '-minion-' ; then
         solver=minion
@@ -25,6 +31,8 @@ else
             MS_GSS_OPTS='--propagate-using-lackey always --send-partials-to-lackey'
         elif echo ${alg} | grep -q -- '-comm-rollback-' ; then
             MS_GSS_OPTS='--propagate-using-lackey root-and-backjump'
+        elif echo ${alg} | grep -q -- '-comm-randomrollback-' ; then
+            MS_GSS_OPTS='--propagate-using-lackey random-and-backjump'
         fi
     fi
 
@@ -69,6 +77,19 @@ started=$(date +'%s.%N' )
 if [[ $solver == gss ]] ; then
     glasgow_subgraph_solver --format lad --timeout ${TIMEOUT} ../../../../${pattern} ../../../../${target} > >(tee gss.out)
     retcode=$?
+elif [[ $solver == pathlad ]] ; then
+    ulimit -s 65536
+    ../../../../pathLAD/main -p ../../../../${pattern} -t ../../../../${target} -v -f -s ${TIMEOUT} > >(tee pathlad.out)
+    retcode=$?
+elif [[ $solver == vf2 ]] ; then
+    ../../../../vflib/solve_vf ../../../../${pattern} ../../../../${target} ${TIMEOUT} > >(tee solve_vf.out)
+    retcode=$?
+elif [[ $solver == ri ]] ; then
+    timeout ${TIMEOUT}s ../../../../ri/solve_ri3 mono loopgfu \
+        <(../../../../ri/lad2gfu < ../../../../${target} ) \
+        <(../../../../ri/lad2gfu < ../../../../${pattern} ) > >(tee solve_ri.out)
+    retcode=$?
+    [[ $retcode == 124 ]] && retcode=0
 else
     cp ../../../../code/si-${problem}.essence . || exit 1
     ../../../ladtoessence ../../../../${pattern} ../../../../${target} > instance.param || exit 1

### 2020-11-17 paper.tex
More plottery

@@
 numbers=none,
-keywordstyle=\bfseries,  %\color{red}
+keywordstyle=\bfseries\color{blue},
 keywords = {language, Essence, given, letting, find, such, that,
@@
 
-% \usepackage{showframe}
+\usepackage{showframe}
 
@@
     \includegraphics{gen-graph-nosideconstraints.pdf}\hfill\includegraphics{gen-graph-oddeven.pdf}
-    \medskip
+    \bigskip
     \includegraphics{gen-graph-mostlyodd.pdf}\hfill\includegraphics{gen-graph-parity.pdf}

### 2020-12-05 experiments/run-experiment.sh
More runner options

@@ -12,6 +12,9 @@ MS_SR_OPTS=
 MS_GSS_OPTS=
 if echo ${alg} | grep -q -- '-gss-' ; then
     solver=gss
+    if echo ${alg} | grep -q -- '-parallel-' ; then
+        MS_GSS_OPTS="--parallel"
+    fi
 elif echo ${alg} | grep -q -- '-pathlad'- ; then
     solver=pathlad
 elif echo ${alg} | grep -q -- '-vf2'- ; then
@@ -24,15 +27,15 @@ else
     elif echo ${alg} | grep -q -- '-hybrid-' ; then
         solver=hybrid
         if echo ${alg} | grep -q -- '-comm-checker-' ; then
-            MS_GSS_OPTS='--propagate-using-lackey never'
+            MS_GSS_OPTS="--propagate-using-lackey never $MS_GSS_OPTS"
         elif echo ${alg} | grep -q -- '-comm-test-' ; then
-            MS_GSS_OPTS='--propagate-using-lackey never --send-partials-to-lackey'
+            MS_GSS_OPTS="--propagate-using-lackey never --send-partials-to-lackey $MS_GSS_OPTS"
         elif echo ${alg} | grep -q -- '-comm-propagate-' ; then
-            MS_GSS_OPTS='--propagate-using-lackey always --send-partials-to-lackey'
+            MS_GSS_OPTS="--propagate-using-lackey always --send-partials-to-lackey $MS_GSS_OPTS"
         elif echo ${alg} | grep -q -- '-comm-rollback-' ; then
-            MS_GSS_OPTS='--propagate-using-lackey root-and-backjump'
+            MS_GSS_OPTS="--propagate-using-lackey root-and-backjump $MS_GSS_OPTS"
         elif echo ${alg} | grep -q -- '-comm-randomrollback-' ; then
-            MS_GSS_OPTS='--propagate-using-lackey random-and-backjump'
+            MS_GSS_OPTS="--propagate-using-lackey random-and-backjump $MS_GSS_OPTS"
         fi
     fi
 
@@ -75,7 +78,12 @@ kill_descendants ()
 retcode=x
 started=$(date +'%s.%N' )
 if [[ $solver == gss ]] ; then
-    glasgow_subgraph_solver --format lad --timeout ${TIMEOUT} ../../../../${pattern} ../../../../${target} > >(tee gss.out)
+    GSS_PROBLEM_OPTIONS=
[…]

### 2020-12-05 paper.tex
More words and plots

@@
 
-\title{Subgraph Isomorphism With Side Constraints}
+\title{Finding Subgraphs With Side Constraints}
 \author{
@@
     Jefferson\inst{1} \and Ciaran McCreesh\inst{2} \and Patrick
-    Prosser\inst{2} \and Steffen Zschaler\inst{3} \\
+    Prosser\inst{3} \and Steffen Zschaler\inst{4} \\
 }
@@
     University of Glasgow, Scotland \\ \email{ciaran.mccreesh@glasgow.ac.uk} \and
+    Algorithmicists Anonymous (Machrihanish), Scotland \and
     King's College London, England

### 2020-12-17 paper/paper.tex
Move the paper into a directory

--- 2020-12-05 paper.tex
@@
 
-\usepackage{showframe}
+%\usepackage{showframe}
 
-\title{Finding Subgraphs With Side Constraints}
+\title{Finding Subgraphs With Side Constraints\thanks{This research was supported by the
+Engineering and Physical Sciences Research Council [grant number EP/P026842/1]}}
+
 \author{
@@
     University of Glasgow, Scotland \\ \email{ciaran.mccreesh@glasgow.ac.uk} \and
-    Algorithmicists Anonymous (Machrihanish), Scotland \and
+    Algorithmicists Anonymous, Scotland \and
     King's College London, England
@@
 
-Finding or counting small ``pattern'' graphs inside larger ``target'' graphs is a widely applicable
-hard problem, with applications including ??. This has led to the development of numerous dedicated
+Finding or counting small ``pattern'' graphs inside larger ``target'' graphs are widely applicable
+hard problems, with applications including ??. This has led to the development of numerous dedicated
 algorithms, with the Glasgow Subgraph Solver \cite{DBLP:conf/gg/McCreeshP020} being the current
@@
 hybrid modelling system can be driven by the Essence high level modelling language
-\cite{DBLP:journals/constraints/FrischHJHM08}, making it accessible to non-specialists.
+\cite{DBLP:journals/constraints/FrischHJHM08} and the Conjure toolchain, making it accessible to
+non-specialists.
 
@@
 preserved.  It is also common to want to count or enumerate all solutions, rather than deciding
-whether at least one solution exists.  Subsets of these variations are all supported by many
+whether at least one solution exists.  Subsets of these variations are supported by many
 dedicated subgraph isomorphism algorithms, including the Glasgow Subgraph Solver.
@@
 \begin{lstlisting}
-such that forAll (a, b) in defined(pat) .
-    pat((a, b)) = tgt((f(a), f(b)))
+such that forAll ((a, b), lbl) in pat .
+    ((f(a), f(b)), lbl) in toSet(tgt)
 \end{lstlisting}
@@
 \end{lstlisting}
-And the Conjure tool tells us that there are exactly two solutions to the problem, as we would
-expect:
+Using the Conjure tool to compile Essence to a constraint programming model which is then solved by
+Minion, we find there are exactly two solutions to the problem, as we would expect:
 \begin{lstlisting}
@@
 \end{lstlisting}
-Supporting other problem variants and constraints is similarly straightforward, even if auxiliary
-variables are required. For example, if instead we want to allow relabelling on vertex labels (which
-is typically not supported by dedicated solvers), we could do the following:
-\begin{lstlisting}
-find r : function (total, injective) L --> L
-such that forAll a : P .  r(plab(a)) = tlab(f(a))
-\end{lstlisting}
-and we would find two additional solutions,
-\begin{lstlisting}
-(1 --> 1, 2 --> 3, 3 --> 5)
-(1 --> 2, 2 --> 3, 3 --> 5)
-\end{lstlisting}
-and if we removed the injective keyword for the relabelling, we would find a fifth mapping
-\begin{lstlisting}
-(1 --> 2, 2 --> 3, 3 --> 1)
-\end{lstlisting}
+As we will see in \cref{section:problems}, supporting other problem variants and constraints is
+similarly straightforward, even if auxiliary variables are required. %For example, if instead we
+%want to allow relabelling on vertex labels (which
+%is typically not supported by dedicated solvers), we could do the following:
+%\begin{lstlisting}
+%find r : function (total, injective) L --> L
+%such that forAll a : P .  r(plab(a)) = tlab(f(a))
+%\end{lstlisting}
+%and we would find two additional solutions,
+%\begin{lstlisting}
+%(1 --> 1, 2 --> 3, 3 --> 5)
+%(1 --> 2, 2 --> 3, 3 --> 5)
+%\end{lstlisting}
+%and if we removed the injective keyword for the relabelling, we would find a fifth mapping
+%\begin{lstlisting}
+%(1 --> 2, 2 --> 3, 3 --> 1)
+%\end{lstlisting}
+%We return to relabelling in \cref{section:typegraphs}.
 
-\subsection{Experiments and Motivation}
+\subsection{Preliminary Experiments and Motivation}
 
@@
 Unfortunately, whilst elegant and flexible, the performance of this approach is rather terrible on
-basic subgraph isomorphism instances.  We performed computational experiments on a cluster of
-??fatanodes spec, using the ??14,621 unlabelled, undirected instances from Solnon's benchmark
-suite\footnote{https://perso.liris.cnrs.fr/christine.solnon/SIP.html}. This benchmark suite was
-originally designed for algorithm portfolios work \cite{DBLP:conf/lion/KotthoffMS16}, and brings
-together several collections of application and randomly-generated instances with varying
+basic subgraph isomorphism instances.  The computational experiments in this paper are performed on
+a cluster of machines with dual Intel Xeon E5-2697A v4 processors and 512GBytes RAM running Ubuntu
+18.04.  Initially, we will be using the 14,621 unlabelled, undirected instances from Solnon's
+benchmark suite\footnote{https://perso.liris.cnrs.fr/christine.solnon/SIP.html}. This benchmark
+suite was originally designed for algorithm portfolios work \cite{DBLP:conf/lion/KotthoffMS16}, and
+brings together several collections of application and randomly-generated instances with varying
 difficulties and solution counts (including many unsatisfiable instances). Some of the instances
-have up to ?? vertices and ?? edges in patterns and up to ?? vertices and ?? edges in
-targets; these lead to rather large models, by constraint programming standards.
+have up to 900 vertices and 14,420 edges in patterns and up to 6,671 vertices and 209,000 edges in
+targets. These lead to rather large models, by constraint programming standards: the largest
+generated table constraint has nearly half a million entries.
 
-In \cref{figure:solvers} we plot the cumulative number of instances solved over time for
-the non-induced decision problem, comparing the high level approach to the Glasgow Subgraph Solver
+In \cref{figure:solvers} we plot the cumulative number of instances solved over time for the
+non-induced decision problem, comparing the high level approach to the Glasgow Subgraph Solver
 \cite{DBLP:conf/gg/McCreeshP020} and PathLAD \cite{DBLP:conf/lion/KotthoffMS16} (the two strongest
@@
 The high level approach has very slow startup times (which is to be expected as it involves
-launching a Java virtual machine), but much more worryingly, only just catches up with the worst
-other solver in number of instances solved as the timeout approaches. Worse, as the scatter plot in
-\cref{figure:solvers} shows, there are almost no instances where the high level approach
-does better than the Glasgow Subgraph Solver. And, although not pictured, the situation for the
-induced variant is even more pessimistic.
+launching a Java virtual machine and reading in a very large table constraint), but much more
+worryingly, only catches up with the worst other solver in number of instances solved as the timeout
+approaches. Worse, as the scatter plot in \cref{figure:solvers} shows, there are almost no instances
+where the high level approach does better than the Glasgow Subgraph Solver. (For induced problems,
+the results are even less favourable.)
 
@@
 applications, but we do not want to have to abandon the performance that dedicated solvers can get
-on hard instances. In ??section we evaluate several ways of using a CP solver in
+on hard instances. In \cref{section:hybrid} we evaluate several ways of using a CP solver in
 conjunction with the Glasgow Subgraph Solver, with a focus on low level implementation details. In
-??section we then return to high level modelling, and look at the convenience it provides for
-??temporal problems, ??relabelling and retyping, and ??optimisation problems.
+\cref{section:problems} we then return to high level modelling, and look at the convenience it
+provides for ??temporal problems, ??relabelling and retyping, and ??optimisation problems.
 
-\section{Hybrid Solving}
+\section{Hybrid Solving}\label{section:hybrid}
 
-The Glasgow Subgraph Solver employs a CP approach to solve subgraph-finding problems, but using
-special data structures and algorithms---for example, rather than representing the adjacency
-constraint using a table, it uses bitset adjacency matrices. The solver also exploits various graph
-invariants involving degrees and paths to further reduce the search space, and also has special
-search order heuristics. From this paper's perspective, the most important design aspect is that
-internally, the solver has a CP style variable for each vertex in the pattern graph, whose domains
-range over the vertices of the target graph. The solver performs a backtracking search\footnote{The
-solver also makes use of restarts and nogood recording, rather than just na\"\i{}ve backtracking,
-but this does not hugely affect this paper.}
-attempting to assign each variable a value from its domain, whilst respecting adjacency and
-injectivity constraints. At each recursive call of search, the solver performs \emph{propagation} to
-eliminate infeasible values from domains. If any domain becomes empty, the solver backtracks;
-otherwise, it selects a variable, and tries assigning it each value from its domain in turn.
+The Glasgow Subgraph Solver \cite{DBLP:conf/gg/McCreeshP020} employs a CP approach to solve
+subgraph-finding problems, but using special data structures and algorithms---for example, rather
+than representing the adjacency constraint using a table, it uses bitset adjacency matrices
+\cite{DBLP:conf/cp/McCreeshP15}. The solver also exploits various graph invariants involving degrees
+\cite{DBLP:journals/constraints/ZampelliDS10} and paths \cite{DBLP:conf/cp/AudemardLMGP14} to
+further reduce the search space, and employs special search order heuristics
+\cite{DBLP:conf/cpaior/ArchibaldDHMP019}. From this paper's perspective, the most important design
+aspect is that internally, the solver has a CP style variable for each vertex in the pattern graph,
+whose domains range over the vertices of the target graph. The solver performs a backtracking search
+with restarts and nogood recording, attempting to assign each variable a value from its domain,
+whilst respecting adjacency and injectivity constraints. At each recursive call of search, the
+solver performs \emph{propagation} to eliminate infeasible values from domains. If any domain
+becomes empty, the solver backtracks; otherwise, it selects a variable, and tries assigning it each
+value from its domain in turn.
 
@@
 to form a correspondence between their internal representations. We will allow the CP solver to have
-additional variables that the subgraph solver does not know about (such as the second mapping used
-in the relabelling example in the previous section), and we do not specifically require the CP
-solver to be aware of all of the graph constraints.  This gives us several potential basic
-approaches to handling side constraints:
-\begin{itemize}
-    \item We could use a CP solver as a \emph{solution checker}. Whenever the subgraph solver finds
-        a solution, it will pass it to the CP solver, which will treat the solution as a set of
-        equality constraints. The CP solver will then attempt to find a satisfying assignment. If
-        the CP solver does not have any additional variables, this is equivalent to simply checking
-        that the remaining constraints hold, but otherwise it can involve search. For a decision
-        problem, the CP solver then communicates back to the subgraph solver either ``yes, this is a
-        valid solution'', or ``no, reject this solution and keep going''. If we
-        are solving a counting or enumeration problem, the CP solver must find \emph{all} solutions
-        and communicate this back to the subgraph solver.
-    \item We could additionally ask a CP solver at every stage of search to \emph{test} whether the
-        state the subgraph solver is in is obviously infeasible. Whenever the subgraph solver has
-        finished performing propagation, it can communicate the \emph{trail} (that is, its current
-        sequence of guessed assignments) to the CP solver, which again treats these as additional
-        equality constraints. The CP solver then performs its own propagation (but not search), and
-        communicates back either a ``yes, keep going'' or a ``no, backtrack immediately''.
-    \item Finally, after this testing, we could also ask the CP solver to communicate any deletions
-        it infers back to the subgraph solver. In other words, the subgraph solver would use the CP
-        solver as an additional propagator.
-\end{itemize}
+additional variables that the subgraph solver does not know about, and we do not specifically
+require the CP solver to be aware of all of the graph constraints. We could therefore use a CP
+solver as a \emph{solution checker}. Whenever the subgraph solver finds a solution, it will pass it
+to the CP solver, which will treat the solution as a set of equality constraints. The CP solver will
+then attempt to find a satisfying assignment. If the CP solver does not have any additional
+variables, this is equivalent to simply checking that the remaining constraints hold, but otherwise
+it can involve search. For a decision problem, the CP solver then communicates back to the subgraph
+solver either ``yes, this is a valid solution'', or ``no, reject this solution and keep going''. If
+we are solving a counting or enumeration problem, the CP solver must find \emph{all} solutions and
+communicate this back to the subgraph solver.
+
+For more power, but possibly also greater cost, we could additionally ask a CP solver at every stage
+of search to test whether the state the subgraph solver is in is obviously infeasible.  Whenever the
+subgraph solver has finished performing propagation, it can communicate the \emph{trail} (that is,
+its current sequence of guessed assignments) to the CP solver, which again treats these as
+additional equality constraints. The CP solver then performs its own propagation (but not search),
+and communicates back either a ``yes, keep going'' or a ``no, backtrack immediately''.  Finally,
+after this testing, we could also ask the CP solver to communicate any deletions it infers back to
+the subgraph solver. In other words, the subgraph solver would use the CP solver as an additional
+propagator.
+
 Unfortunately, each of these approaches has drawbacks. The solution we will settle upon is based
@@
 
-?? How we do this: FIFOs, simple text based protocol. Both solvers are run and initialised. The
-subgraph solver calls the CP solver as a propagator. The CP solver sets up the model, then clones it
-and adds in new constraints for each query.
+To enable communication between the two solvers, we use FIFOs (named pipes), and a simple text-based
+protocol.  Both solvers are run and initialised, and then the subgraph solver proceeds as normal,
+whilst the CP solver waits to be given commands. The subgraph solver thens communicate its trail
+or a candidate solution as a set of assignment constraints, which the CP solver reads. The CP solver
+clones its initial model, adds in the given set of constraints, and then either propagates or
+searches as directed. It then communicates its success or failure state, and any deletions, back to
+the subgraph solver. This approach is reasonably solver-agnostic, and adding support for different
+CP solvers (or non-CP solvers) should not be difficult.
 
-\subsection{Experiments}
+?? How does the Essence / Conjure side work?
+
+\subsection{Design Experiments}
 
@@
     \includegraphics{gen-graph-mostlyodd.pdf}\hfill\includegraphics{gen-graph-parity.pdf}
-    \caption{Comparing different approaches to hybrid solving. On the top row, ``no side
-    constraints'' then with the ``more odd target vertices than even target vertices'' side
-    constraint; on the bottom row, the ``mostly odd target vertices'' side constraint on the left,
-    and the ``odd to odd, even to even'' side constraint on the right.}\label{figure:cumulative}
+    \caption{Comparing different approaches to hybrid solving, showing the cumulative number of
+    instances solved over time. On the top row, ``no side constraints'' then with the ``more odd
+    target vertices than even target vertices'' side constraint; on the bottom row, the ``mostly odd
+    target vertices'' side constraint on the left, and the ``odd to odd, even to even'' side
+    constraint on the right. In the top left plot, the ``Checking'' and ``Rollback'' lines are
+    indistinguishable.}\label{figure:cumulative}
 \end{figure}
@@
 We now present the results of some computational experiments. The experiments in this section are
-designed to be hard, and to emphasise the difference between the approaches, rather than to be
-realistic. We will continue to work with Solnon's non-induced subgraph isomorphism benchmark
-instances, but will consider three variations. Firstly, we will consider the problem with no side
-constraints. This is, in some sense, the worst case scenario, where we must pay the full price of
-hybrid solving, but cannot get any benefit from it. Secondly, let's say that the number of odd
-target vertices used must be greater than the number of even target vertices used:
+designed to be computationally hard, and to emphasise the difference between the approaches, rather
+than to be realistic. We will continue to work with Solnon's non-induced subgraph isomorphism
+benchmark instances, but will consider three variations. Firstly, we will consider the problem with
+no side constraints. This is, in some sense, the worst case scenario, where we must pay the full
+price of hybrid solving, but cannot get any benefit from it. Secondly, let us say that the number of
+odd-indexed target vertices used must be greater than the number of even target vertices used:
 \newcommand{\modsymbol}{\%}
@@
 \end{lstlisting}
-Thirdly, let us say that fewer than three odd target vertices may be used:
+Thirdly, let us instead say that fewer than three odd target vertices may be used:
 \begin{lstlisting}
@@
 
-We chose these last three problems because intuitively, the first is likely not to be able to
+We chose these last three problems because the first is likely not to be able to
 perform inference until deep in a search tree (when most pattern vertices are mapped to specific
 target vertices), the second is likely to be able to perform inference early in the search tree
-(after a few assignments have been made, but not at the root node), and the third will propagate
+(after a few assignments have been made, but not at the root node), and the third should propagate
 only at the root node.
 
-The results of these experiments are presented in
-\cref{figure:cumulative}. Let us first look at the top left plot, where we do not actually have any side constraints. When calling
-the CP solver as a solution checker, we ultimately achieve the same performance as the Glasgow
-Subgraph Solver\footnote{Actually, because high level modelling can use different names for
-variables and values, we get slight differences due to changes to tiebreaking in search order
-heuristics.}, although we can pay a substantial startup overhead. This should not be surprising:
-with no side constraints, the subgraph solver runs as normal, and will perform just one call to the
-CP solver on satisfiable instances. The testing and propagating approaches are both over an order of
-magnitude slower: calling the CP solver at every search node is clearly very expensive---although if
-we do call the CP solver, having it communicate deletions is not much of an additional overhead.
-Finally, all approaches substantially outperform using a CP solver on its own without help from the
-subgraph solver.
+The results of these experiments are presented in \cref{figure:cumulative}. Let us first look at the
+top left plot, where we do not actually have any side constraints. When calling the CP solver as a
+solution checker, we ultimately achieve the same performance as the Glasgow Subgraph
+Solver\footnote{Actually, because high level modelling can use different names for variables and
+values, we get slight differences due to changes to tiebreaking in search order heuristics.},
+although we can pay a substantial startup overhead. This should not be surprising: with no side
+constraints, the subgraph solver runs as normal, and will perform just one call to the CP solver on
+satisfiable instances. The propagating approach is over an order of magnitude slower: calling the CP
+solver at every search node is clearly very expensive. (We also tried calling the CP solver to test
+feasibility, without communicating deletions; this made no noticable difference to performance, and
+so is not pictured.) Finally, all approaches substantially outperform using a CP solver on its own
+without help from the subgraph solver.
 
-?? Include a version where Minion doesn't actually get the constraints
+What about the remaining three plots in \cref{figure:cumulative}, where we do have side constraints?
+As we hoped, we see differences between the three plots. On the top right, where we expect the side
+constraints to fire late, solution checking clearly beats propagation during search. However, on the
+bottom left, where we expect side constraints to fire early, the propagating approach is much better
+than solution checking. Meanwhile, on the bottom right, where our constraints fire only at the root
+node, checking performs extremely poorly compared to propagating. In each case, any hybrid approach using
+the subgraph solver remains much better than a pure CP approach, except that in bottom right plot
+checking is slightly worse than just using the CP solver on its own.
 
-What about the remaining three plots in \cref{figure:cumulative}, where we do have side constraints (and so can no
-longer compare to the subgraph solver on its own)? As we hoped, we see differences between the three
-plots. On the top right, where we expect the side constraints to fire late, solution
-checking clearly beats testing or propagation during search. However, on the bottom left,
-where we expect side constraints to fire early, the testing and propagating approaches are much
-better than solution checking. ?? Bottom right. Finally, any approach using the subgraph solver
-remains much better than a pure CP approach.
+As a point of comparison, we also implemented these three sets of side constraints natively inside
+the Glasgow solver. For the ``more odd than even'' case, we did this through solution checking, for
+``less than three odd'' we implemented checking during search, and for ``odd to odd, even to even''
+we implemented initial domain filtering. Our implementation choices here are intended to reflect what a
+reasonable programmer would do if the high level approach were not available (and we intentionally
+selected side constraints that would not be too difficult to implement). In two of the three cases,
+the hand-crafted code does somewhat outperform the hybrid solving, but in the ``more odd than even''
+case, hybrid solving actually beats the hand-crafted dedicated solver implementation when using the
+rollback approach, which we now describe.
 
-\subsection{A Rollback Approach}
+\subsection{A Rollback Approach to Communication}
 
@@
 backtrack. At this point, we call the CP solver again, with full propagation. Either the CP solver
-is happy, in which case we proceed with search, or the CP solver indicates failure, in which case we
-backtrack again, and do another attempt at full propagation, and so on until the CP solver is happy.
+indicates feasibility, in which case we proceed with search (potentially with a reduced set of
+domains), or the CP solver indicates failure, in which case we backtrack again, and do another
+attempt at full propagation, and so on until feasiblity is reached.
 
@@
 
-The results in \cref{figure:nosideconstraints,figure:oddeventhreeodd} demonstrate the
-success of this approach. When there are no side constraints, this approach has no overheads
-compared to solution checking. When constraints fire late, this approach is slightly better than
-solution checking, and when constraints fire early, this approach is slightly better than always
-testing or propagating during search. In other words, rolling back from failures gives us all of the
-strengths and none of the weaknesses of the simpler approaches.
+The results in \cref{figure:cumulative} demonstrate the success of this approach. When there are no
+side constraints, this approach has no overheads compared to solution checking. When constraints
+fire late, this approach is better than solution checking, and when constraints fire early, this
+approach is better than always propagating during search. In other words, rolling back from failures
+gives us all of the strengths and none of the weaknesses of the simpler approaches. We will
+therefore use this method for the remainder of the paper.
 
-?? Randomly also calling in unsat trees?
+\section{Subgraph Problems with Side Constraints}\label{section:problems}
 
-\section{Subgraph Problems with Side Constraints}
+We now look at three classes of real-world subgraph-finding problems that, until now, have been
+solved using dedicated approaches. We show how easy it is to model these problems in Essence. ?? do
+some actual simple experiments on them?
 
@@
 \begin{lstlisting}
-such that forAll (a,b) in defined(pat) .
-    pat((a,b)) = tgt((f(a), f(b)))
+such that forAll ((a, b), lbl) in pat .
+    ((f(a), f(b)), lbl) in toSet(tgt)
 such that forAll a : P .
@@
 \end{lstlisting}
+?? Should we say that we are now searching for a commuting square?
 
@@
 \end{lstlisting}
-Recall that with vertex relabelling, four solutions exist. However, with retyping, only two of the
-solutions respect the type graph,
+We now find two solutions:
 \begin{lstlisting}
@@
 \end{lstlisting}
-whilst the solutions that instead map pattern vertex 3 to target vertex 4 are no longer valid.
+because we can map pattern vertex 3 to target vertex 5 through retyping, but mapping pattern vertex 3
+to target vertex 4 would not respect the type graph structure.
 
@@
 labels that must match exactly. If we look at \cref{figure:littlegraphs}, now ignoring vertex labels
-but using the edge labels,
+but using the edge labels to carry the timestamps,
 \begin{lstlisting}
@@
 
-\subsection{Compiler Instruction Generation}
+Of course, when using a high level modelling approach, we are not restricted to these three problem
+variants, and could easily try out new models in an interactive setting. For example, it would take
+only a few minutes to write a model for a temporal problem where all edges must occur within a short
+but unspecified time period, whereas adapting a dedicated solver to check this constraint would be a
+substantial programming effort (and making the solver propagate rather than check this constraint
+would be even harder).
 
@@
 
-We can also support optimisation problems. If, for example, each target vertex has a cost associated with it,
+The system we created also support optimisation problems (and does not require that the subgraph
+isomorphism solver be aware that this is what is going on). If, for example, each target vertex has
+a cost associated with it,
 \begin{lstlisting}
@@
 \end{lstlisting}
+We could also just as easily ask for the solution whose most expensive edge is cheapest, or that
+uses fewest vertices with a particular label. These kinds of problem occur widely in practice,
+including in ?? skyline graphs, ?? labelled subgraph finding, ?? weighted cliques.
 
 \section{Conclusion}
+
+?? A lot of flexibility, decent performance on hard instances although the startup costs are
+vicious. A key benefit is supporting large queries, which current solutions struggle with. Suitable
+for interactive use in domain settings. For integration into larger systems, would benefit from
+bypassing the high level modelling language and using two low level solvers directly.
 

### 2021-01-13 paper/paper.tex
Two paragraphs on morphisms between typed graphs

@@
 \usetikzlibrary{shapes, positioning, decorations.text}
-\usepackage{cleveref}
+
+\usepackage{todonotes}
 
@@
 }
+
+\usepackage{hyperref}
+\usepackage{cleveref}
 
@@
 
-A \emph{typed graph} is a kind of labelled graph where the labels also carry a graph structure
-??cite. In general, mappings between typed graphs must preserve the label structure, but here we
-will be looking at a problem variation involving relabelling which occurs in ??application. We may
-describe typed graph subisomorphism problems in Essence as follows. As before, we are given a
+  In graph theory~\cite{Ehrig+06}, a graph is given by a set of vertices and a set of edges and two functions giving the source and target node for an edge, respectively.
+  Morphisms between graphs map vertices to vertices and edges to edges, preserving the structure of the graphs.
+  Often, we are interested in typed graphs.
+  Here, we assume a dedicated graph TG---the type graph---to be given.
+  Then we can define typed graphs as graphs for which a morphism into TG has been defined.
+  In effect, we are labelling each vertex and edge with its corresponding type.
+
+  For typed graphs, morphisms between the graphs are typically defined between graphs typed over the same type graph, but there are situations where we are interested in mapping between graphs typed over different type graphs.
+  We describe an example scenario in~\cite{Duran+12,Duran+17}, where we compose graph transformation systems by defining morphisms between the rules constituting the respective transformation systems.
+  In this case, the source and target transformation systems will normally have different type graphs and we will first establish a morphism between the two type graphs.
+  Mappings between the various graphs making up the rules then need to preserve structure and typing subject to the morphism between type graphs.
+  We have developed tool support for this approach to specification composition.\footnote{\url{https://github.com/gts-morpher/gts_morpher}}
+  A key objective is to minimise the amount of specification that needs to be written.
+  For example, the tool allows morphisms between graph transformation systems to be only partially specified and then automatically completes the full morphism, if it can do so unambiguously.
+  This requires solving a sub-isomorphism problem.
+
+%A \emph{typed graph} is a kind of labelled graph where the labels also carry a graph structure
+%??cite. In general, mappings between typed graphs must preserve the label structure, but here we
+%will be looking at a problem variation involving relabelling which occurs in ??application. 
+We may describe typed graph subisomorphism problems in Essence as follows. As before, we are given a
 pattern graph and a target graph, both of which carry labels; we will draw the vertex labels from

### 2021-01-14 paper/paper.tex
Adding Jess sentences and cites

@@
 \subsection{Temporal Subgraph Problems}\label{section:temporalgraphs}
+In a \emph{temporal graph}, edges are labelled with timestamps that denote times when edges are active---here we use
+integers as timestamps.  Including information on the timing of edges substantially increases the modelling power of these graphs, allowing them to 
+more accurately reflect the structure and dynamics of a wide variety of real-world systems (e.g. trade networks, changing contact networks, transport networks), 
+and address optimisation questions in which the timing of edges is fundamental.
 
-In a \emph{temporal graph}, edges are labelled with some kind of timestamp---here we will use
-integers. Temporal pattern matching and counting has applications in ??. There are at least three
+As algorithms and formalisms have become available for temporal graphs, examples of their application have become widespread \cite{holmeReview}, 
+notably including applications within epidemiology \cite{temporalEpi} and computational social science \cite{socialTemporalCite}.
+Because the use of temporal graphs has spread beyond theoretical researchers, the ability to rapidly define and experiment with 
+new problem definitions and constraints is valuable - practitioners are unlikely to define bespoke algorithms for novel problems as they arise.
+
+%Temporal graphs 
+% why being able to try new constraints easily is useful for practitioners
+
+%In a \emph{temporal graph}, edges are labelled with some kind of timestamp---here we will use
+%integers. Temporal pattern matching and counting has applications in ??. 
+There are at least three
 common kinds of temporal subgraph isomorphism. In an \emph{exact} subisomorphism, times are simply

### 2021-01-21 paper/paper.tex
Abstract

@@
     theories'' solver that is both performant and flexible. We also show how such an approach can be
-    driven by the Essence high level modelling language. We give practical examples taken from
-    ??applications.
+    driven by the Essence high level modelling language, giving ease of modelling to non-expert
+    users. We give practical examples involving
+    temporal graphs, typed graphs from software engineering, and costed subgraph isomorphism
+    problems.
 \end{abstract}

### 2021-01-25 paper/paper.tex
the Essence bit

@@
 
-?? How does the Essence / Conjure side work?
+Problems specified in Essence are converted to concrete models that can be solved by a CP solver like Minion using Conjure and Savile Row.
+Running Conjure with a special flag (--graph-solver) instructs it to generate an extra file for the consumption of the graph solver, in addition to a CP model written in Essence Prime.
+The model is then converted to Minion input. The graph solver file includes the mappings between the nodes and the edges in the graph to the names of the decision variables used in the model.
 

### 2021-01-25 paper/paper.tex
Add implementation section

--- 2021-01-21 paper/paper.tex
@@
 then attempt to find a satisfying assignment. If the CP solver does not have any additional
-variables, this is equivalent to simply checking that the remaining constraints hold, but otherwise
-it can involve search. For a decision problem, the CP solver then communicates back to the subgraph
+variables, this is equivalent to simply checking that the remaining constraints hold, but in general
+this will require search.
+For a decision problem, the CP solver then communicates back to the subgraph
 solver either ``yes, this is a valid solution'', or ``no, reject this solution and keep going''. If
@@
 
+Problems are specified in Essence and then are converted to concrete models that can be solved by
+a CP solver (in our experiments Minion) using Conjure and Savile Row.
+Running Conjure with a special flag (--graph-solver) instructs it to generate an extra file which
+describes how the variables representing the graph is represented in the SavileRow input (known
+as Essence'). This is used
+by the graph solver so it can map between its internal state and the variables in Essence'.
+SavileRow converts the graph model to Minion input, and this conversion contains information which
+allows Minion to map its internal state back to the Essence' given to SavileRow. The graph solver
+and Minion then communicate mappings between the nodes and edges in the graph using the
+identifiers in the Essence' representation of the problem.
+
 To enable communication between the two solvers, we use FIFOs (named pipes), and a simple text-based
 protocol.  Both solvers are run and initialised, and then the subgraph solver proceeds as normal,
-whilst the CP solver waits to be given commands. The subgraph solver thens communicate its trail
-or a candidate solution as a set of assignment constraints, which the CP solver reads. The CP solver
-clones its initial model, adds in the given set of constraints, and then either propagates or
-searches as directed. It then communicates its success or failure state, and any deletions, back to
-the subgraph solver. This approach is reasonably solver-agnostic, and adding support for different
-CP solvers (or non-CP solvers) should not be difficult.
-
-?? How does the Essence / Conjure side work?
+whilst the CP solver waits to be given commands. The subgraph solver then communicates its trail
+or a candidate solution as a set of assignment constraints to the CP solver. The CP solver
+then sets these assignments its state and then either performs a single propagation, or
+complete search, as directed. When finished the solver communicates its success or failure state,
+and any deletions (if requested), back to the subgraph solver and reverts any changes made by
+setting the assignments. This approach is designed to be
+solver-agnostic, and adding support for different CP solvers (or non-CP solvers) is simple, as long
+as they support performing a search or propagation from a given set of assignments.
 

### 2021-01-27 paper/paper.tex
More or less done

@@
 \author{
-    \"Ozg\"ur Akg\"un\inst{1} \and Jessica Enright\inst{2} \and Christopher
-    Jefferson\inst{1} \and Ciaran McCreesh\inst{2} \and Patrick
-    Prosser\inst{3} \and Steffen Zschaler\inst{4} \\
+    \"Ozg\"ur Akg\"un\inst{1}\orcidID{0000-0003-4460-6912} \and Jessica
+    Enright\inst{2}\orcidID{0000-0002-0266-3292} \and Christopher
+    Jefferson\inst{1}\orcidID{0000-0003-2979-5989} \and Ciaran
+    McCreesh\inst{2}\orcidID{0000-0002-6106-4871} \and Patrick
+    Prosser\inst{3}\orcidID{0000-0003-4460-6912} \and Steffen
+    Zschaler\inst{4}\orcidID{0000-0001-9062-6637} \\
 }
@@
     graph. There are excellent dedicated solvers for this problem, but they require substantial
-    programming effort to handle complex side constraints; however, general purpose constraint
-    solvers struggle on more difficult graph instances. We show how to combine the state of the art
-    Glasgow Subgraph Solver with the Minion constraint programming solver to get a ``subgraphs modulo
-    theories'' solver that is both performant and flexible. We also show how such an approach can be
-    driven by the Essence high level modelling language, giving ease of modelling to non-expert
-    users. We give practical examples involving
-    temporal graphs, typed graphs from software engineering, and costed subgraph isomorphism
-    problems.
+    programming effort to handle the complex side constraints that often occur in practical
+    applications of the problem; however, general purpose constraint solvers struggle on more
+    difficult graph instances. We show how to combine the state of the art Glasgow Subgraph Solver
+    with the Minion constraint programming solver to get a ``subgraphs modulo theories'' solver that
+    is both performant and flexible. We also show how such an approach can be driven by the Essence
+    high level modelling language, giving ease of modelling and prototyping to non-expert users. We
+    give practical examples involving temporal graphs, typed graphs from software engineering, and
+    costed subgraph isomorphism problems.
 \end{abstract}
@@
 
-Finding or counting small ``pattern'' graphs inside larger ``target'' graphs are widely applicable
-hard problems, with applications including ??. This has led to the development of numerous dedicated
+Finding small ``pattern'' graphs inside larger ``target'' graphs is a widely applicable
+hard problem, with applications including compilers \cite{DBLP:conf/cp/BlindellLCS15},
+bioinformatics \cite{DBLP:journals/bmcbi/BonniciGPSF13,Elhesha2021}, chemistry \cite{o:Regin95},
+malware detection \cite{DBLP:conf/dimva/BruschiMM06}, pattern recognition
+\cite{DBLP:journals/ijprai/FoggiaPV14}, and the design of mechanical locks
+\cite{DBLP:journals/siamrev/VomelLBF17}.  This has led to the development of numerous dedicated
 algorithms, with the Glasgow Subgraph Solver \cite{DBLP:conf/gg/McCreeshP020} being the current
@@
 require either extensive programming or inefficient post-processing. This paper explores a different
-approach: by allowing the Glasgow Subgraph Solver to use the Minion constraint programming (CP) solver
-\cite{DBLP:conf/ecai/GentJM06} for side constraints, we achieve both the performance only a
-dedicated solver can offer, with the flexibility of a full CP toolkit. This
-hybrid modelling system can be driven by the Essence high level modelling language
+approach: by allowing the Glasgow Subgraph Solver to use the Minion constraint programming (CP)
+solver \cite{DBLP:conf/ecai/GentJM06} for side constraints, we achieve both the performance only a
+dedicated solver can offer, with the flexibility of a full CP toolkit. This hybrid modelling system
+can be driven by the Essence high level modelling language
 \cite{DBLP:journals/constraints/FrischHJHM08} and the Conjure toolchain, making it accessible to
@@
 As we will see in \cref{section:problems}, supporting other problem variants and constraints is
-similarly straightforward, even if auxiliary variables are required. %For example, if instead we
-%want to allow relabelling on vertex labels (which
-%is typically not supported by dedicated solvers), we could do the following:
-%\begin{lstlisting}
-%find r : function (total, injective) L --> L
-%such that forAll a : P .  r(plab(a)) = tlab(f(a))
-%\end{lstlisting}
-%and we would find two additional solutions,
-%\begin{lstlisting}
-%(1 --> 1, 2 --> 3, 3 --> 5)
-%(1 --> 2, 2 --> 3, 3 --> 5)
-%\end{lstlisting}
-%and if we removed the injective keyword for the relabelling, we would find a fifth mapping
-%\begin{lstlisting}
-%(1 --> 2, 2 --> 3, 3 --> 1)
-%\end{lstlisting}
-%We return to relabelling in \cref{section:typegraphs}.
+similarly straightforward, even if auxiliary variables are required. For example, if instead we want
+to allow relabelling on vertex labels (which is typically not supported by dedicated solvers), we
+could do the following:
+\begin{lstlisting}
+find r : function (total, injective) L --> L
+such that forAll a : P .  r(plab(a)) = tlab(f(a))
+\end{lstlisting}
+and we would find two additional solutions,
+\begin{lstlisting}
+(1 --> 1, 2 --> 3, 3 --> 5)
+(1 --> 2, 2 --> 3, 3 --> 5)
+\end{lstlisting}
+and if we removed the injective keyword for the relabelling, we would find a fifth mapping
+\begin{lstlisting}
+(1 --> 2, 2 --> 3, 3 --> 1)
+\end{lstlisting}
+We return to relabelling in \cref{section:typegraphs}.
 
 \subsection{Preliminary Experiments and Motivation}
+
+Unfortunately, whilst elegant and flexible, the performance of this approach leaves a lot to be
+desired on basic subgraph isomorphism instances.  The computational experiments in this paper are
+performed on a cluster of machines with dual Intel Xeon E5-2697A v4 processors and 512GBytes RAM
+running Ubuntu 18.04.  We will be using the 14,621 unlabelled, undirected instances from
+Solnon's benchmark suite\footnote{https://perso.liris.cnrs.fr/christine.solnon/SIP.html}. This
+benchmark suite was originally designed for algorithm portfolios work
+\cite{DBLP:conf/lion/KotthoffMS16}, and brings together several collections of application and
+randomly-generated instances with varying difficulties and solution counts (including many
+unsatisfiable instances). Some of the instances have up to 900 vertices and 14,420 edges in patterns
+and up to 6,671 vertices and 209,000 edges in targets. These lead to rather large models, by
+constraint programming standards: the largest generated table constraint has nearly half a million
+entries. However, these sizes are realistic from an applications perspective, and it would be
+desirable if solvers could handle even larger target graphs.
 
@@
 
-Unfortunately, whilst elegant and flexible, the performance of this approach is rather terrible on
-basic subgraph isomorphism instances.  The computational experiments in this paper are performed on
-a cluster of machines with dual Intel Xeon E5-2697A v4 processors and 512GBytes RAM running Ubuntu
-18.04.  Initially, we will be using the 14,621 unlabelled, undirected instances from Solnon's
-benchmark suite\footnote{https://perso.liris.cnrs.fr/christine.solnon/SIP.html}. This benchmark
-suite was originally designed for algorithm portfolios work \cite{DBLP:conf/lion/KotthoffMS16}, and
-brings together several collections of application and randomly-generated instances with varying
-difficulties and solution counts (including many unsatisfiable instances). Some of the instances
-have up to 900 vertices and 14,420 edges in patterns and up to 6,671 vertices and 209,000 edges in
-targets. These lead to rather large models, by constraint programming standards: the largest
-generated table constraint has nearly half a million entries.
-
 In \cref{figure:solvers} we plot the cumulative number of instances solved over time for the
@@
 \cite{DBLP:conf/gg/McCreeshP020} and PathLAD \cite{DBLP:conf/lion/KotthoffMS16} (the two strongest
-CP approaches), and to VF2 \cite{DBLP:journals/pami/CordellaFSV04} and RI
+CP-inspired approaches), and to VF2 \cite{DBLP:journals/pami/CordellaFSV04} and RI
 \cite{DBLP:journals/bmcbi/BonniciGPSF13} (simpler algorithms which perform well on easy instances).
@@
 \cref{section:problems} we then return to high level modelling, and look at the convenience it
-provides for ??temporal problems, ??relabelling and retyping, and ??optimisation problems.
+provides for retyping typed graphs, for temporal problems, and for optimisation problems.
 
-\section{Hybrid Solving}\label{section:hybrid}
+\section{Hybrid Solving With High-Level Modelling}\label{section:hybrid}
 
-The Glasgow Subgraph Solver \cite{DBLP:conf/gg/McCreeshP020} employs a CP approach to solve
-subgraph-finding problems, but using special data structures and algorithms---for example, rather
-than representing the adjacency constraint using a table, it uses bitset adjacency matrices
+Designing an effective hybrid solving system involved three major decisions: how the high-level
+modelling language would identify suitable problems to hybridise, how the solvers would communicate,
+and how often this communication would occur. The first two decisions were relatively
+straightforward to make, but the third required using computational experiments to evaluate
+different options. This section discusses all three of these decisions.
+
+\subsection{High-Level Modelling}
+
+We chose to use the Essence high-level constraint modelling language
+\cite{DBLP:journals/constraints/FrischHJHM08} because of its support for convenient high-level types
+like functions and relations. In a conventional modelling pipeline, problems are specified in
+Essence and then are converted to concrete models that can be solved by a CP solver (in our
+experiments Minion) via the Conjure and SavileRow tools. We augmented Conjure with a command line
+option that instructs it to generate an extra file which describes how the variables representing
+the graph are represented in the SavileRow input (known as Essence'). This is used by the graph
+solver so it can map between its internal state and the variables in Essence'.  SavileRow converts
+the graph model to Minion input, and this conversion contains information which allows Minion to map
+its internal state back to the Essence' given to SavileRow. The graph solver and Minion then
+communicate mappings between the nodes and edges in the graph using the identifiers in the Essence'
+representation of the problem.
+
+This design is based upon the notion that a CP solver and a subgraph solver can have enough of a
+shared understanding of a problem to solve it co-operatively. Indeed, the Glasgow Subgraph Solver
+\cite{DBLP:conf/gg/McCreeshP020} employs a CP approach to solve subgraph-finding problems, but using
+special data structures and algorithms---for example, rather than representing the adjacency
+constraint using an explicit table, it uses bitset adjacency matrices
 \cite{DBLP:conf/cp/McCreeshP15}. The solver also exploits various graph invariants involving degrees
@@
 
-For now we will assume that there is some way of setting up the subgraph solver and a CP solver such
-that they both have this same set of variables and values, and that there is some way of knowing how
-to form a correspondence between their internal representations. We will allow the CP solver to have
-additional variables that the subgraph solver does not know about, and we do not specifically
-require the CP solver to be aware of all of the graph constraints. We could therefore use a CP
-solver as a \emph{solution checker}. Whenever the subgraph solver finds a solution, it will pass it
-to the CP solver, which will treat the solution as a set of equality constraints. The CP solver will
-then attempt to find a satisfying assignment. If the CP solver does not have any additional
-variables, this is equivalent to simply checking that the remaining constraints hold, but in general
-this will require search.
-For a decision problem, the CP solver then communicates back to the subgraph
-solver either ``yes, this is a valid solution'', or ``no, reject this solution and keep going''. If
-we are solving a counting or enumeration problem, the CP solver must find \emph{all} solutions and
-communicate this back to the subgraph solver.
+The high-level approach, then, gives us a way of setting up the subgraph solver and a CP solver such
+that they both have an equivalent same set of variables and values for the graph part of the model,
+and tells us how to form a correspondence between their internal representations. Importantly, this
+allows the CP solver to have additional variables that the subgraph solver does not know about, and
+we do not specifically require the CP solver to be aware of all of the graph constraints.
+(Additionally, due to preprocessing, the CP solver may also sometimes have only a subset of values
+for some graph variables visible to it.)
+
+\subsection{When to Communicate?}
+
+Having found a way to set up the two solvers, we must next ask when they should communicate. The
+simplest approach would be to use the CP solver as a \emph{solution checker} for the subgraph
+solver.  Whenever the subgraph solver finds a solution, it will pass it to the CP solver, which will
+treat the solution as a set of equality constraints. The CP solver will then attempt to find a
+satisfying assignment. If the CP solver does not have any additional variables, this is equivalent
+to simply checking that the remaining constraints hold, but in general this will require search.
+For a decision problem, the CP solver then communicates back to the subgraph solver either ``yes,
+this is a valid solution'', or ``no, reject this solution and keep going''. If we are solving a
+counting or enumeration problem, the CP solver must find \emph{all} solutions and communicate this
+back to the subgraph solver.
 
 For more power, but possibly also greater cost, we could additionally ask a CP solver at every stage
-of search to test whether the state the subgraph solver is in is obviously infeasible.  Whenever the
+of search to test whether the subgraph solver is in an obviously infeasible state.  Whenever the
 subgraph solver has finished performing propagation, it can communicate the \emph{trail} (that is,
@@
 
-\subsection{Implementation}
-
-Problems are specified in Essence and then are converted to concrete models that can be solved by
-a CP solver (in our experiments Minion) using Conjure and Savile Row.
-Running Conjure with a special flag (--graph-solver) instructs it to generate an extra file which
-describes how the variables representing the graph is represented in the SavileRow input (known
-as Essence'). This is used
-by the graph solver so it can map between its internal state and the variables in Essence'.
-SavileRow converts the graph model to Minion input, and this conversion contains information which
-allows Minion to map its internal state back to the Essence' given to SavileRow. The graph solver
-and Minion then communicate mappings between the nodes and edges in the graph using the
-identifiers in the Essence' representation of the problem.
+\subsection{How to Communicate}
 
@@
 protocol.  Both solvers are run and initialised, and then the subgraph solver proceeds as normal,
-whilst the CP solver waits to be given commands. The subgraph solver then communicates its trail
-or a candidate solution as a set of assignment constraints to the CP solver. The CP solver
-then sets these assignments its state and then either performs a single propagation, or
-complete search, as directed. When finished the solver communicates its success or failure state,
-and any deletions (if requested), back to the subgraph solver and reverts any changes made by
-setting the assignments. This approach is designed to be
-solver-agnostic, and adding support for different CP solvers (or non-CP solvers) is simple, as long
-as they support performing a search or propagation from a given set of assignments.
+whilst the CP solver waits to be given commands. The subgraph solver then communicates its trail or
+a candidate solution as a set of assignment constraints to the CP solver. The CP solver then sets
+these assignments its state and then either performs a single propagation, or complete search, as
+directed. When finished the solver communicates its success or failure state, and any deletions (if
+requested), back to the subgraph solver and reverts any changes made by setting the assignments.
+This approach is designed to be solver-agnostic, and adding support for different CP solvers (or
+non-CP solvers) is simple, as long as they support performing a search or propagation from a given
+set of assignments.
 
@@
 We now present the results of some computational experiments. The experiments in this section are
-designed to be computationally hard, and to emphasise the difference between the approaches, rather
+designed to be hard, and to emphasise the difference between the approaches, rather
 than to be realistic. We will continue to work with Solnon's non-induced subgraph isomorphism
-benchmark instances, but will consider three variations. Firstly, we will consider the problem with
+benchmark instances, but will consider four variations. Firstly, we will consider the problem with
 no side constraints. This is, in some sense, the worst case scenario, where we must pay the full
@@
 
-We chose these last three problems because the first is likely not to be able to
+We chose these problems because the second is likely not to be able to
 perform inference until deep in a search tree (when most pattern vertices are mapped to specific
-target vertices), the second is likely to be able to perform inference early in the search tree
-(after a few assignments have been made, but not at the root node), and the third should propagate
+target vertices), the third is likely to be able to perform inference early in the search tree
+(after a few assignments have been made, but not at the root node), and the fourth should propagate
 only at the root node.
@@
 solver at every search node is clearly very expensive. (We also tried calling the CP solver to test
-feasibility, without communicating deletions; this made no noticable difference to performance, and
+feasibility, without communicating deletions; this made no noticeable difference to performance, and
 so is not pictured.) Finally, all approaches substantially outperform using a CP solver on its own
@@
 As a point of comparison, we also implemented these three sets of side constraints natively inside
-the Glasgow solver. For the ``more odd than even'' case, we did this through solution checking, for
-``less than three odd'' we implemented checking during search, and for ``odd to odd, even to even''
+the Glasgow solver. For the ``more odd than even'' case, we did this through solution checking; for
+``less than three odd'' we implemented checking during search; and for ``odd to odd, even to even''
 we implemented initial domain filtering. Our implementation choices here are intended to reflect what a
@@
 time during search, but too expensive to call it \emph{all} of the time. We will therefore introduce
-a new approach, which we call \emph{rollback}. This approach is inspired by backjumping ??cite, as
-well as by the conflict analysis methods used in SAT and SMT solvers ??cite and in lazy clause
-generating CP solvers ??cite. The idea is as follows.  Firstly, we call the CP solver with full
-propagation at the root node, in case we are dealing with a particularly rich labelling scheme.
-Secondly, we use the CP solver for solution checking, since this is required for correctness. Now,
-suppose the CP solver rejects a candidate solution: this will cause the subgraph solver to
-backtrack. At this point, we call the CP solver again, with full propagation. Either the CP solver
-indicates feasibility, in which case we proceed with search (potentially with a reduced set of
-domains), or the CP solver indicates failure, in which case we backtrack again, and do another
-attempt at full propagation, and so on until feasiblity is reached.
+a new approach, which we call \emph{rollback}. This approach is inspired by backjumping
+\cite{DBLP:journals/ci/Prosser93}, as well as by the conflict analysis methods used in SAT and SMT
+solvers \cite{DBLP:conf/iccad/SilvaS96} and in lazy clause generating CP solvers
+\cite{DBLP:conf/cp/OhrimenkoSC07,DBLP:conf/cpaior/Stuckey10}. The idea is as follows.  Firstly, we
+call the CP solver with full propagation at the root node, in case we are dealing with a
+particularly rich labelling scheme.  Secondly, we use the CP solver for solution checking, since
+this is required for correctness. Now, suppose the CP solver rejects a candidate solution: this will
+cause the subgraph solver to backtrack. At this point, we call the CP solver again, with full
+propagation. Either the CP solver indicates feasibility, in which case we proceed with search
+(potentially with a reduced set of domains), or the CP solver indicates failure, in which case we
+backtrack again, and do another attempt at full propagation, and so on until feasibility is reached.
 
@@
 We now look at three classes of real-world subgraph-finding problems that, until now, have been
-solved using dedicated approaches. We show how easy it is to model these problems in Essence. ?? do
-some actual simple experiments on them?
+solved using dedicated approaches. We show how easy it is to model these problems in Essence,
+demonstrating the usefulness of the high-level modelling approach for prototyping and development.
 
@@
 
-  In graph theory~\cite{Ehrig+06}, a graph is given by a set of vertices and a set of edges and two functions giving the source and target node for an edge, respectively.
-  Morphisms between graphs map vertices to vertices and edges to edges, preserving the structure of the graphs.
-  Often, we are interested in typed graphs.
-  Here, we assume a dedicated graph TG---the type graph---to be given.
-  Then we can define typed graphs as graphs for which a morphism into TG has been defined.
-  In effect, we are labelling each vertex and edge with its corresponding type.
+The basic notion of a graph conveys only adjacency information, and a subgraph isomorphism simply
+finds a certain structural pattern. In practice, this is often augmented with additional
+information---for example, we have seen how labels can be associated with vertices and edges, which
+can be used in chemical applications to represent different kinds of atom or bond. In this case,
+subgraph isomorphisms are also expected to preserve labels, so carbon atoms can only be mapped to
+carbon atoms, and double bonds must be mapped to double bonds.  A richer labelling abstraction comes
+in the form of \emph{typed graphs}, where the labels themselves also carry a graph structure
+\cite{Ehrig+06}; we show an example in \cref{figure:littlegraphs}. In practice this labelling
+structure is specified by providing two graphs, together with a morphism from the main graph to its
+type graph.
 
-  For typed graphs, morphisms between the graphs are typically defined between graphs typed over the same type graph, but there are situations where we are interested in mapping between graphs typed over different type graphs.
-  We describe an example scenario in~\cite{Duran+12,Duran+17}, where we compose graph transformation systems by defining morphisms between the rules constituting the respective transformation systems.
-  In this case, the source and target transformation systems will normally have different type graphs and we will first establish a morphism between the two type graphs.
-  Mappings between the various graphs making up the rules then need to preserve structure and typing subject to the morphism between type graphs.
-  We have developed tool support for this approach to specification composition.\footnote{\url{https://github.com/gts-morpher/gts_morpher}}
-  A key objective is to minimise the amount of specification that needs to be written.
-  For example, the tool allows morphisms between graph transformation systems to be only partially specified and then automatically completes the full morphism, if it can do so unambiguously.
-  This requires solving a sub-isomorphism problem.
+For typed graphs, morphisms between the graphs are typically defined between graphs typed over the
+same type graph, but there are situations where we are interested in mapping between graphs typed
+over different type graphs.  One such scenario from a software engineering context is described by
+Dur{\'a}n et al.~\cite{Duran+12,Duran+17}, where graph transformation systems are composed by
+defining morphisms between the rules constituting the respective transformation systems.  In this
+case, the source and target transformation systems will normally have different type graphs, a
+morphism must also be established between the two type graphs.  Mappings between the various graphs
+making up the rules then need to preserve structure and typing subject to the morphism between type
+graphs. This approach to specification composition is implemented by the gts-morpher
+tool.\footnote{\url{https://github.com/gts-morpher/gts_morpher}} A key objective is to minimise the
+amount of specification that needs to be written.  For example, the tool allows morphisms between
+graph transformation systems to be only partially specified and then automatically completes the
+full morphism, if it can do so unambiguously---this requires solving a sub-isomorphism problem.
 
-%A \emph{typed graph} is a kind of labelled graph where the labels also carry a graph structure
-%??cite. In general, mappings between typed graphs must preserve the label structure, but here we
-%will be looking at a problem variation involving relabelling which occurs in ??application. 
 We may describe typed graph subisomorphism problems in Essence as follows. As before, we are given a
@@
 \end{lstlisting}
-?? Should we say that we are now searching for a commuting square?
 
@@
 
+More generally, the field of model-driven software engineering includes numerous examples of using
+search and optimisation techniques to generate or transform graphs
+\cite{DBLP:journals/ase/BoussaidSA17}. Existing approaches largely make use of ad-hoc
+\cite{DBLP:conf/icse/SemerathNV18} and metaheuristic methods
+\cite{DBLP:conf/models/EfstathiouWZ14,DBLP:journals/sosym/BillFTMW19,DBLP:conf/models/BurduselZ019},
+but we believe that with the help of suitably accessible high-level modelling tools, this could
+become a fruitful area for constraint programming research in the future.
+
 \subsection{Temporal Subgraph Problems}\label{section:temporalgraphs}
-In a \emph{temporal graph}, edges are labelled with timestamps that denote times when edges are active---here we use
-integers as timestamps.  Including information on the timing of edges substantially increases the modelling power of these graphs, allowing them to 
-more accurately reflect the structure and dynamics of a wide variety of real-world systems (e.g. trade networks, changing contact networks, transport networks), 
-and address optimisation questions in which the timing of edges is fundamental.
 
-As algorithms and formalisms have become available for temporal graphs, examples of their application have become widespread \cite{holmeReview}, 
-notably including applications within epidemiology \cite{temporalEpi} and computational social science \cite{socialTemporalCite}.
-Because the use of temporal graphs has spread beyond theoretical researchers, the ability to rapidly define and experiment with 
-new problem definitions and constraints is valuable - practitioners are unlikely to define bespoke algorithms for novel problems as they arise.
+Another labelling scheme is used in \emph{temporal graphs}, where edges are labelled with timestamps
+that denote times when edges are active---here we use integers as timestamps.  Including information
+on the timing of edges substantially increases the modelling power of these graphs, allowing them to
+more accurately reflect the structure and dynamics of a wide variety of real-world systems (e.g.
+trade networks, changing contact networks, transport networks), and address optimisation questions
+in which the timing of edges is fundamental.
 
-%Temporal graphs 
-% why being able to try new constraints easily is useful for practitioners
+As algorithms and formalisms have become available for temporal graphs, examples of their
+application have become widespread \cite{holmeReview}, notably including applications within
+epidemiology \cite{temporalEpi} and computational social science \cite{socialTemporalCite}.  Because
+the use of temporal graphs has spread beyond theoretical researchers, the ability to rapidly define
+and experiment with new problem definitions and constraints is valuable---practitioners are unlikely
+to define bespoke algorithms for novel problems as they arise.
 
-%In a \emph{temporal graph}, edges are labelled with some kind of timestamp---here we will use
-%integers. Temporal pattern matching and counting has applications in ??. 
-There are at least three
-common kinds of temporal subgraph isomorphism. In an \emph{exact} subisomorphism, times are simply
-labels that must match exactly. If we look at \cref{figure:littlegraphs}, now ignoring vertex labels
-but using the edge labels to carry the timestamps,
+There are at least three common kinds of temporal subgraph isomorphism. In an \emph{exact}
+subisomorphism, times are simply labels that must match exactly. If we look at
+\cref{figure:littlegraphs}, now ignoring vertex labels but using the edge labels to carry the
+timestamps,
 \begin{lstlisting}
@@
 only a few minutes to write a model for a temporal problem where all edges must occur within a short
-but unspecified time period, whereas adapting a dedicated solver to check this constraint would be a
-substantial programming effort (and making the solver propagate rather than check this constraint
-would be even harder).
+but unspecified time period \cite{DBLP:conf/asunam/RedmondC13}, whereas adapting a dedicated solver
+to check this constraint would be a substantial programming effort (and making the solver propagate
+rather than check this constraint would be even harder).
 
@@
 uses fewest vertices with a particular label. These kinds of problem occur widely in practice,
-including in ?? skyline graphs, ?? labelled subgraph finding, ?? weighted cliques.
+including in skyline graph queries \cite{DBLP:journals/pvldb/PandeR017}, labelled subgraph finding
+\cite{DBLP:conf/colognetwente/DellOlmoCC11}, and weighted clique problems
+\cite{DBLP:conf/cp/McCreeshPST17}.
 
@@
 
-?? A lot of flexibility, decent performance on hard instances although the startup costs are
-vicious. A key benefit is supporting large queries, which current solutions struggle with. Suitable
-for interactive use in domain settings. For integration into larger systems, would benefit from
-bypassing the high level modelling language and using two low level solvers directly.
+The system we have presented shows that it is possible to combine the power of modern subgraph
+solvers with the flexibility of a general purpose constraint programming toolkit, although doing so
+efficiently requires careful consideration of how frequently the solvers communicate. We believe
+further research in this direction may be useful---for example, would it be possible to make use of
+some kind of conflict analysis rather than a backjumping approach?
+
+When driven by a high level modelling approach, this system is particularly suitable for rapid
+prototyping and for dynamic queries where side constraints can be specified in response to user
+need. However, the high level modelling approach does come with a large startup cost, which makes it
+unsuitable for deployment in application contexts that involve solving many thousands of problem
+instances in real-time. Fortunately though, connecting the low level solvers manually is also an
+option once a design has been decided upon. We also expect that new approaches may be necessary to
+deal with the huge but sparse graphs that arise in some applications, since table constraints and
+conventional CP domain stores both struggle when moving beyond ten thousand of vertices in target
+graphs.
 

### 2021-01-27 paper/paper.tex
Minor fixes

@@
 defining morphisms between the rules constituting the respective transformation systems.  In this
-case, the source and target transformation systems will normally have different type graphs, a
+case, the source and target transformation systems will normally have different type graphs; a
 morphism must also be established between the two type graphs.  Mappings between the various graphs
 making up the rules then need to preserve structure and typing subject to the morphism between type
-graphs. This approach to specification composition is implemented by the gts-morpher
+graphs. This approach to specification composition is implemented by the GTSMorpher
 tool.\footnote{\url{https://github.com/gts-morpher/gts_morpher}} A key objective is to minimise the

### 2021-01-28 paper/paper.tex
Final changes

@@
 }
-\authorrunning{\"O. Akgun et al.}
+\authorrunning{\"O. Akg\"un et al.}
 \institute{
@@
 
-We can express these problems in the Essence high level modelling language, as follows. We assume
+We can express these problems in the Essence modelling language, as follows. We assume
 vertices take their labels from the set $L = \{ 1\ldots\ell \}$ for some given $\ell$, and edges
@@
 \end{lstlisting}
-We take as input a pattern directed graph which has $p$ vertices (which we number from 1 to $p$, in
-the set $P$), and a target directed graph which has $t$ vertices (numbered from 1 to $t$, the set
+We take as input a directed pattern graph which has $p$ vertices (which we number from 1 to $p$, in
+the set $P$), and a directed target graph which has $t$ vertices (numbered from 1 to $t$, the set
 $T$). Each graph is represented as total function from vertices to vertex labels, and a
@@
 
-We chose to use the Essence high-level constraint modelling language
-\cite{DBLP:journals/constraints/FrischHJHM08} because of its support for convenient high-level types
-like functions and relations. In a conventional modelling pipeline, problems are specified in
-Essence and then are converted to concrete models that can be solved by a CP solver (in our
-experiments Minion) via the Conjure and SavileRow tools. We augmented Conjure with a command line
-option that instructs it to generate an extra file which describes how the variables representing
-the graph are represented in the SavileRow input (known as Essence'). This is used by the graph
-solver so it can map between its internal state and the variables in Essence'.  SavileRow converts
-the graph model to Minion input, and this conversion contains information which allows Minion to map
-its internal state back to the Essence' given to SavileRow. The graph solver and Minion then
-communicate mappings between the nodes and edges in the graph using the identifiers in the Essence'
-representation of the problem.
+We chose to use the Essence modelling language \cite{DBLP:journals/constraints/FrischHJHM08} because
+of its support for convenient high-level types like functions and relations, which can easily
+describe graphs and related abstractions. In a conventional modelling pipeline, problems are
+specified in Essence and then are converted to concrete models that can be solved by a CP solver (in
+our experiments Minion) via the Conjure and SavileRow tools. We augmented Conjure with a command
+line option that instructs it to generate an extra file which describes how the variables
+representing the graph are represented in the SavileRow input (known as Essence'). This is used by
+the graph solver so it can map between its internal state and the variables in Essence'.  SavileRow
+converts the graph model to Minion input, and this conversion contains information which allows
+Minion to map its internal state back to the Essence' given to SavileRow. The graph solver and
+Minion then communicate mappings between the nodes and edges in the graph using the identifiers in
+the Essence' representation of the problem.
 
@@
 The high-level approach, then, gives us a way of setting up the subgraph solver and a CP solver such
-that they both have an equivalent same set of variables and values for the graph part of the model,
+that they both have an equivalent set of variables and values for the graph part of the model,
 and tells us how to form a correspondence between their internal representations. Importantly, this
@@
 a candidate solution as a set of assignment constraints to the CP solver. The CP solver then sets
-these assignments its state and then either performs a single propagation, or complete search, as
+these assignments as its state and either performs a single propagation, or complete search, as
 directed. When finished the solver communicates its success or failure state, and any deletions (if
@@
     \includegraphics{gen-graph-nosideconstraints.pdf}\hfill\includegraphics{gen-graph-oddeven.pdf}
+
     \bigskip
+
     \includegraphics{gen-graph-mostlyodd.pdf}\hfill\includegraphics{gen-graph-parity.pdf}
@@
     constraint on the right. In the top left plot, the ``Checking'' and ``Rollback'' lines are
-    indistinguishable.}\label{figure:cumulative}
+    indistinguishable. The Glasgow+ lines show the Glasgow Subgraph Solver with
+    manually-implemented side constraints.}\label{figure:cumulative}
 \end{figure}
@@
 As a point of comparison, we also implemented these three sets of side constraints natively inside
-the Glasgow solver. For the ``more odd than even'' case, we did this through solution checking; for
-``less than three odd'' we implemented checking during search; and for ``odd to odd, even to even''
-we implemented initial domain filtering. Our implementation choices here are intended to reflect what a
-reasonable programmer would do if the high level approach were not available (and we intentionally
-selected side constraints that would not be too difficult to implement). In two of the three cases,
-the hand-crafted code does somewhat outperform the hybrid solving, but in the ``more odd than even''
-case, hybrid solving actually beats the hand-crafted dedicated solver implementation when using the
-rollback approach, which we now describe.
+the Glasgow solver: we label these as Glasgow+. For the ``more odd than even'' case, we did this
+through solution checking; for ``less than three odd'' we implemented checking during search; and
+for ``odd to odd, even to even'' we implemented initial domain filtering. Our implementation choices
+here are intended to reflect what a reasonable programmer would do if the high level approach were
+not available (and we intentionally selected side constraints that would not be too difficult to
+implement). In two of the three cases, the hand-crafted code does somewhat outperform the hybrid
+solving, but in the ``more odd than even'' case, hybrid solving actually beats the hand-crafted
+dedicated solver implementation when using the rollback approach, which we now describe.
 
@@
 
-A less strict kind of subisomorphism is an \emph{offset}, where edge labels must match exactly but
+A less strict kind of subisomorphism is an \emph{offset}, where edge labels must match exactly, but
 offset by an integer constant. In our example, this means ``find a mapping where the event from 2

### 2021-04-20 paper/paper.tex
Camera ready

@@
     McCreesh\inst{2}\orcidID{0000-0002-6106-4871} \and Patrick
-    Prosser\inst{3}\orcidID{0000-0003-4460-6912} \and Steffen
-    Zschaler\inst{4}\orcidID{0000-0001-9062-6637} \\
+    Prosser\inst{2}\orcidID{0000-0003-4460-6912} \and Steffen
+    Zschaler\inst{3}\orcidID{0000-0001-9062-6637} \\
 }
@@
     University of Glasgow, Scotland \\ \email{ciaran.mccreesh@glasgow.ac.uk} \and
-    Algorithmicists Anonymous, Scotland \and
     King's College London, England
@@
 
-\subsection{Preliminary Experiments and Motivation}
+\subsection{Initial Experiments and Motivation}
 
@@
 performed on a cluster of machines with dual Intel Xeon E5-2697A v4 processors and 512GBytes RAM
-running Ubuntu 18.04.  We will be using the 14,621 unlabelled, undirected instances from
-Solnon's benchmark suite\footnote{https://perso.liris.cnrs.fr/christine.solnon/SIP.html}. This
+running Ubuntu 18.04. The source code used for these experiments is released as part of the Glasgow
+Subgraph
+Solver\footnote{\url{https://github.com/ciaranm/glasgow-subgraph-solver/releases/tag/cpaior2021-finding-subgraphs-with-side-constraints}},
+Minion\footnote{\url{https://github.com/minion/minion/releases/tag/1.9}}, and
+Conjure\footnote{\url{https://github.com/conjure-cp/conjure}} distributions.
+
+For graphs, we will be using the 14,621 unlabelled, undirected instances from
+Solnon's benchmark suite\footnote{\url{https://perso.liris.cnrs.fr/christine.solnon/SIP.html}}. This
 benchmark suite was originally designed for algorithm portfolios work

### 2021-04-20 paper/paper.tex
Better line breaks, and code

@@
 \usepackage{cleveref}
+
+% url by default does horrible things with line breaks, this makes it behave better
+\usepackage{xurl}
 
@@
 Minion\footnote{\url{https://github.com/minion/minion/releases/tag/1.9}}, and
-Conjure\footnote{\url{https://github.com/conjure-cp/conjure}} distributions.
+Conjure\footnote{\url{https://github.com/conjure-cp/conjure}} distributions, and we provide a
+separate archive for experimental
+scripts\footnote{\url{https://github.com/ciaranm/cpaior2021-finding-subgraphs-with-side-constraints}}.
 

### 2021-06-30 talk/talk.tex
Talk

% vim: set spell spelllang=en tw=100 et sw=4 sts=4 foldmethod=marker foldmarker={{{,}}} :

\documentclass{beamer}

\usepackage{tikz}
\usepackage{xcolor}
\usepackage{complexity}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{amsmath}                   % \operatorname
\usepackage{amsfonts}                  % \mathcal
\usepackage{amssymb}                   % \nexists
\usepackage[vlined]{algorithm2e} % algorithms
\usepackage{centernot}
\usepackage{listings}
\usepackage{csquotes}
\usepackage{fancyvrb}

\RequirePackage[tt=false, type1=true]{libertine}
\RequirePackage[varqu]{zi4}
\RequirePackage[libertine]{newtxmath}
\RequirePackage[T1]{fontenc}

\newcommand*{\rom}[1]{\emph{\romannumeral #1 \relax}}

\usetikzlibrary{shapes, arrows, shadows, calc, positioning, fit}
\usetikzlibrary{decorations.pathreplacing, decorations.pathmorphing, shapes.misc}
\usetikzlibrary{tikzmark, backgrounds}
\usetikzlibrary{trees}

\tikzset{processarrow/.style={->, very thick, decorate, decoration={snake, post length=0.5mm}}}
\tikzset{brace/.style={decorate, decoration={brace}, very thick}}

\definecolor{uofguniversityblue}{rgb}{0, 0.219608, 0.396078}

\definecolor{uofgheather}{rgb}{0.356863, 0.32549, 0.490196}
\definecolor{uofgaquamarine}{rgb}{0.603922, 0.72549, 0.678431}
\definecolor{uofgslate}{rgb}{0.309804, 0.34902, 0.380392}
\definecolor{uofgrose}{rgb}{0.823529, 0.470588, 0.709804}
\definecolor{uofgmocha}{rgb}{0.709804, 0.564706, 0.47451}
\definecolor{uofgsandstone}{rgb}{0.321569, 0.278431, 0.231373}
\definecolor{uofgforest}{rgb}{0, 0.2, 0.129412}
\definecolor{uofglawn}{rgb}{0.517647, 0.741176, 0}
\definecolor{uofgcobalt}{rgb}{0, 0.615686, 0.92549}
\definecolor{uofgturquoise}{rgb}{0, 0.709804, 0.819608}
\definecolor{uofgsunshine}{rgb}{1.0, 0.862745, 0.211765}
\definecolor{uofgpumpkin}{rgb}{1.0, 0.72549, 0.282353}
\definecolor{uofgthistle}{rgb}{0.584314, 0.070588, 0.447059}
\definecolor{uofgrust}{rgb}{0.603922, 0.227451, 0.023529}
\definecolor{uofgburgundy}{rgb}{0.490196, 0.133333, 0.223529}
\definecolor{uofgpillarbox}{rgb}{0.701961, 0.047059, 0}
\definecolor{uofglavendar}{rgb}{0.356863, 0.301961, 0.580392}

% {{{ theme things
\useoutertheme[footline=authortitle]{miniframes}
\useinnertheme{rectangles}

\setbeamerfont{block title}{size={}}
\setbeamerfont{title}{size=\large,series=\bfseries}
\setbeamerfont{section title}{size=\large,series=\mdseries}
\setbeamerfont{author}{size=\normalsize,series=\mdseries}
\setbeamercolor*{structure}{fg=uofguniversityblue}
\setbeamercolor*{palette primary}{use=structure,fg=black,bg=white}
\setbeamercolor*{palette secondary}{use=structure,fg=white,bg=uofgcobalt}
\setbeamercolor*{palette tertiary}{use=structure,fg=white,bg=uofguniversityblue}
\setbeamercolor*{palette quaternary}{fg=white,bg=black}

\setbeamercolor*{titlelike}{parent=palette primary}

\beamertemplatenavigationsymbolsempty

\setbeamertemplate{title page}
{
    \begin{tikzpicture}[remember picture, overlay]
        \node at (current page.north west) {
            \begin{tikzpicture}[remember picture, overlay]
                \fill [fill=uofguniversityblue, anchor=north west] (0, 0) rectangle (\paperwidth, -3.0cm);
            \end{tikzpicture}
        };

        \node (logo) [anchor=north east, shift={(-2.2cm,-0.4cm)}] at (current page.north east) {
            \includegraphics[keepaspectratio=true,scale=0.4]{UoG_keyline.pdf}
        };
        \node (logo2) [anchor=north, below = 0 of logo.south] {
            \includegraphics[keepaspectratio=true,scale=0.025]{UStA_Logo_PR-REV.png}
        };
        \node (logo3) [anchor=west, right = 0 of $(logo2.east)!0.5!(logo.east)$] {
            \includegraphics[keepaspectratio=true,scale=0.1]{kings.png}
        };

        \node [anchor=west, xshift=0.1cm, yshift=-0.6cm] at (current page.west |- logo.west) {
            \begin{minipage}{0.65\paperwidth}\raggedright
                {\usebeamerfont{title}\usebeamercolor[white]{}\inserttitle}\\[0.1cm]
                {\usebeamerfont{author}\usebeamercolor[white]{}\insertauthor}
            \end{minipage}
        };
    \end{tikzpicture}
}

\setbeamertemplate{section page}
{
    \begin{centering}
        \begin{beamercolorbox}[sep=12pt,center]{part title}
            \usebeamerfont{section title}\insertsection\par
        \end{beamercolorbox}
    \end{centering}
}

\newcommand{\frameofframes}{/}
\newcommand{\setframeofframes}[1]{\renewcommand{\frameofframes}{#1}}

\makeatletter
\setbeamertemplate{footline}
{%
    \begin{beamercolorbox}[colsep=1.5pt]{upper separation line foot}
    \end{beamercolorbox}
    \begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,%
        leftskip=.3cm,rightskip=.3cm plus1fil]{author in head/foot}%
        \leavevmode{\usebeamerfont{author in head/foot}\insertshortauthor}%
        \hfill%
        {\usebeamerfont{institute in head/foot}\usebeamercolor[fg]{institute in head/foot}\insertshortinstitute}%
    \end{beamercolorbox}%
    \begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,%
        leftskip=.3cm,rightskip=.3cm plus1fil]{title in head/foot}%
        {\usebeamerfont{title in head/foot}\insertshorttitle}%
        \hfill%
        {\usebeamerfont{frame number}\usebeamercolor[fg]{frame number}\insertframenumber~\frameofframes~\inserttotalframenumber}
    \end{beamercolorbox}%
    \begin{beamercolorbox}[colsep=1.5pt]{lower separation line foot}
    \end{beamercolorbox}
}

\makeatletter
\newenvironment{nearlyplainframe}[2][]{
    \def\beamer@entrycode{\vspace*{-\headheight}\vspace*{3pt}}
    \setbeamertemplate{headline}
    {%
        \begin{beamercolorbox}[colsep=1.5pt]{upper separation line head}
        \end{beamercolorbox}
        \begin{beamercolorbox}[ht=0.5ex,dp=0.125ex,%
            leftskip=.3cm,rightskip=.3cm plus1fil]{title in head/foot}%
        \end{beamercolorbox}%
        \begin{beamercolorbox}[ht=0.5ex,dp=0.125ex,%
            leftskip=.3cm,rightskip=.3cm plus1fil]{author in head/foot}%
        \end{beamercolorbox}%
        \begin{beamercolorbox}[colsep=1.5pt]{lower separation line head}
        \end{beamercolorbox}
        \vspace*{\headheight}
    }

    \setbeamertemplate{footline}
    {%
        \begin{beamercolorbox}[colsep=1.5pt]{upper separation line foot}
        \end{beamercolorbox}
        \begin{beamercolorbox}[ht=0.5ex,dp=0.125ex,%
            leftskip=.3cm,rightskip=.3cm plus1fil]{author in head/foot}%
        \end{beamercolorbox}%
        \begin{beamercolorbox}[ht=0.5ex,dp=0.125ex,%
            leftskip=.3cm,rightskip=.3cm plus1fil]{title in head/foot}%
        \end{beamercolorbox}%
        \begin{beamercolorbox}[colsep=1.5pt]{lower separation line foot}
        \end{beamercolorbox}
    }

    \begin{frame}[#1]{#2}
    }{
    \end{frame}
}
\makeatother

\makeatletter
\newenvironment{justborderframe}[2][]{
    \def\beamer@entrycode{\vspace*{-\headheight}}
    \setbeamertemplate{headline}
    {%
        \begin{beamercolorbox}[colsep=1.5pt]{upper separation line head}
        \end{beamercolorbox}
        \begin{beamercolorbox}[ht=0.5ex,dp=0.125ex,%
            leftskip=.3cm,rightskip=.3cm plus1fil]{title in head/foot}%
        \end{beamercolorbox}%
        \begin{beamercolorbox}[ht=0.5ex,dp=0.125ex,%
            leftskip=.3cm,rightskip=.3cm plus1fil]{author in head/foot}%
        \end{beamercolorbox}%
        \begin{beamercolorbox}[colsep=1.5pt]{lower separation line head}
        \end{beamercolorbox}
        \vspace*{\headheight}
    }

    \setbeamertemplate{footline}
    {%
        \begin{beamercolorbox}[colsep=1.5pt]{upper separation line foot}
        \end{beamercolorbox}
        \begin{beamercolorbox}[ht=0.5ex,dp=0.125ex,%
            leftskip=.3cm,rightskip=.3cm plus1fil]{author in head/foot}%
        \end{beamercolorbox}%
        \begin{beamercolorbox}[ht=0.5ex,dp=0.125ex,%
            leftskip=.3cm,rightskip=.3cm plus1fil]{title in head/foot}%
        \end{beamercolorbox}%
        \begin{beamercolorbox}[colsep=1.5pt]{lower separation line foot}
        \end{beamercolorbox}
    }

    \begin{frame}[#1]{}
    }{
    \end{frame}
}
\makeatother

% }}}

\author[\"Ozg\"ur Akg\"un \and Jessica Enright \and Christopher Jefferson \and \textbf{Ciaran
McCreesh} \and Patrick Prosser \and Steffen Zschaler]{
    \"Ozg\"ur Akg\"un \and Jessica Enright \and \\
    Christopher Jefferson \and Ciaran McCreesh \and \\
    Patrick Prosser \and Steffen Zschaler
}
\title[Finding Subgraphs With Side Constraints]{Finding Subgraphs \\With Side Constraints}

\begin{document}

{
    \usebackgroundtemplate{
        \tikz[overlay, remember picture]
        \node[at=(current page.south), anchor=south, inner sep=0pt]{\includegraphics[keepaspectratio=true, height=\paperheight]{background.jpg}};
    }
    \begin{frame}[plain,noframenumbering]
        \titlepage
    \end{frame}
}

\begin{frame}{Modelling and Solving Subgraph Isomorphism}
    \begin{minipage}{0.25\paperwidth}
        \begin{tikzpicture}
        \node [draw, fill] (P1) {}; \node [above=0cm of P1, font=\scriptsize] { 1 };
        \node [draw, fill, right=1cm of P1] (P2) {}; \node [above=0cm of P2, font=\scriptsize] { 2 };
        \node [draw, fill, circle, right=1cm of P2, inner sep=2.5pt] (P3) {}; \node [above=0cm of P3, font=\scriptsize] { 3 };
        \draw [->, shorten >=1mm, shorten <=1mm] (P1) -- (P2); node [midway, font=\scriptsize\itshape, above] { 1 };
        \draw [->, shorten >=1mm, shorten <=1mm] (P2) -- (P3); node [midway, font=\scriptsize\itshape, above] { 2 };
        \node [above=0.7cm of P1.west, anchor=west, xshift=-0.2cm] { pat: };

        \node [draw, fill, below=2cm of P1] (T1) {}; \node [above=0cm of T1, font=\scriptsize] { 1 };
        \node [draw, fill, right=1cm of T1] (T3) {}; \node [above=0cm of T3, font=\scriptsize] { 3 };
        \node [draw, fill, below=0.7cm of T3] (T2) {}; \node [below=0cm of T2, font=\scriptsize] { 2 };
        \node [draw, fill, circle, right=1cm of T3, inner sep=2.5pt] (T4) {}; \node [above=0cm of T4, font=\scriptsize] { 4 };
        \node [draw, fill, diamond, below=0.7cm of T4, inner sep=2pt] (T5) {}; \node [below=0cm of T5, font=\scriptsize] { 5 };
        \draw [->, shorten >=1mm, shorten <=1mm, transform canvas={yshift=0.5mm}] (T1) -- (T3); % node [midway, font=\scriptsize\itshape, above] { 2 };
        \draw [<-, shorten >=1mm, shorten <=1mm, transform canvas={yshift=-0.5mm}] (T1) -- (T3); % node [midway, font=\scriptsize\itshape, below] { 2 };
        \draw [->, shorten >=1mm, shorten <=1mm] (T2) -- (T3); % node [midway, font=\scriptsize\itshape, left] { 1 };
        \draw [->, shorten >=1mm, shorten <=1mm] (T3) -- (T4); % node [midway, font=\scriptsize\itshape, above] { 2 };
        \draw [->, shorten >=1mm, shorten <=1mm] (T3) -- (T5); % node [midway, font=\scriptsize\itshape, below] { 3 };
        \node [above=0.7cm of T1.west, anchor=west, xshift=-0.2cm] { tgt: };

        \node [above=1cm of P1] { ~ };
        \node [below=1cm of T2] { ~ };
            \end{tikzpicture}\end{minipage}\begin{minipage}[c]{0.1\paperwidth}~\end{minipage}\begin{minipage}[c]{0.6\paperwidth}
        \only<2>{\lstinputlisting[basicstyle=\scriptsize\ttfamily]{sip.essence}}%
        \only<3>{\lstinputlisting[basicstyle=\scriptsize\ttfamily]{sip-instance.essence}}%
        \only<4>{\lstinputlisting[basicstyle=\scriptsize\ttfamily]{sip-constraints.essence}}%
        \only<5>{\lstinputlisting[basicstyle=\scriptsize\ttfamily]{sip-solution.essence}}%
        \only<6>{\lstinputlisting[basicstyle=\scriptsize\ttfamily]{sip-induced.essence}}%
        \only<7>{\lstinputlisting[basicstyle=\scriptsize\ttfamily]{sip-induced-solution.essence}}%
    \end{minipage}
\end{frame}

\begin{frame}{Temporal Problems}
    \begin{minipage}{0.25\paperwidth}
        \begin{tikzpicture}
        \node [draw, fill] (P1) {}; \node [above=0cm of P1, font=\scriptsize] { 1 };
        \node [draw, fill, right=1cm of P1] (P2) {}; \node [above=0cm of P2, font=\scriptsize] { 2 };
        \node [draw, fill, right=1cm of P2] (P3) {}; \node [above=0cm of P3, font=\scriptsize] { 3 };
        \draw [->, shorten >=1mm, shorten <=1mm] (P1) -- (P2) node [midway, color=uofgcobalt, font=\scriptsize\itshape, above] { 1 };
        \draw [->, shorten >=1mm, shorten <=1mm] (P2) -- (P3) node [midway, color=uofgcobalt, font=\scriptsize\itshape, above] { 2 };
        \node [above=0.7cm of P1.west, anchor=west, xshift=-0.2cm, font=\scriptsize\itshape] { pat: };

        \node [draw, fill, below=2cm of P1] (T1) {}; \node [above=0cm of T1, font=\scriptsize] { 1 };
        \node [draw, fill, right=1cm of T1] (T3) {}; \node [above=0cm of T3, font=\scriptsize] { 3 };
        \node [draw, fill, below=0.7cm of T3] (T2) {}; \node [below=0cm of T2, font=\scriptsize] { 2 };
        \node [draw, fill, right=1cm of T3] (T4) {}; \node [above=0cm of T4, font=\scriptsize] { 4 };
        \node [draw, fill, below=0.7cm of T4] (T5) {}; \node [below=0cm of T5, font=\scriptsize] { 5 };
        \draw [->, shorten >=1mm, shorten <=1mm, transform canvas={yshift=0.5mm}] (T1) -- (T3) node [midway, color=uofgcobalt, font=\scriptsize\itshape, above] { 2 };
        \draw [<-, shorten >=1mm, shorten <=1mm, transform canvas={yshift=-0.5mm}] (T1) -- (T3) node [midway, color=uofgcobalt, font=\scriptsize\itshape, below] { 2 };
        \draw [->, shorten >=1mm, shorten <=1mm] (T2) -- (T3) node [midway, color=uofgcobalt, font=\scriptsize\itshape, left] { 1 };
        \draw [->, shorten >=1mm, shorten <=1mm] (T3) -- (T4) node [midway, color=uofgcobalt, font=\scriptsize\itshape, above] { 2 };
        \draw [->, shorten >=1mm, shorten <=1mm] (T3) -- (T5) node [midway, color=uofgcobalt, font=\scriptsize\itshape, below] { 3 };
        \node [above=0.7cm of T1.west, anchor=west, xshift=-0.2cm, font=\scriptsize\itshape] { tgt: };

        \node [above=1cm of P1] { ~ };
        \node [below=1cm of T2] { ~ };
            \end{tikzpicture}\end{minipage}\begin{minipage}[c]{0.1\paperwidth}~\end{minipage}\begin{minipage}[c]{0.6\paperwidth}
        \only<2>{\lstinputlisting[basicstyle=\scriptsize\ttfamily]{tsip-instance.essence}}%
        \only<3>{\lstinputlisting[basicstyle=\scriptsize\ttfamily]{tsip-exact-constraints.essence}}%
        \only<4>{\lstinputlisting[basicstyle=\scriptsize\ttfamily]{tsip-exact-solution.essence}}%
        \only<5>{\lstinputlisting[basicstyle=\scriptsize\ttfamily]{tsip-offset-constraints.essence}}%
        \only<6>{\lstinputlisting[basicstyle=\scriptsize\ttfamily]{tsip-offset-solution.essence}}%
        \only<7>{\lstinputlisting[basicstyle=\scriptsize\ttfamily]{tsip-embedding-constraints.essence}}%
        \only<8>{\lstinputlisting[basicstyle=\scriptsize\ttfamily]{tsip-embedding-solution.essence}}%
    \end{minipage}
\end{frame}

\begin{frame}{Retyping Problems}
    \begin{minipage}{0.25\paperwidth}
    \centering\begin{tikzpicture}
        \node [draw, fill] (P1) {}; \node [above=0cm of P1, font=\scriptsize] { 1 };
        \node [draw, fill, right=1cm of P1] (P2) {}; \node [above=0cm of P2, font=\scriptsize] { 2 };
        \node [draw, fill, circle, right=1cm of P2, inner sep=2.5pt] (P3) {}; \node [above=0cm of P3, font=\scriptsize] { 3 };
        \draw [->, shorten >=1mm, shorten <=1mm] (P1) -- (P2);
        \draw [->, shorten >=1mm, shorten <=1mm] (P2) -- (P3);
        \node [above=0.7cm of P1.west, anchor=west, xshift=-0.2cm, font=\scriptsize\itshape] { pat: };

        \begin{scope}[color=uofgcobalt]
        \node [draw, fill, below=1cm of P1] (PT1) {};
        \node [draw, fill, circle, inner sep=2.5pt, right=0.6cm of PT1] (PT2) {};
        \draw [<->, shorten >=1mm, shorten <=1mm] (PT1) -- (PT2);
        \draw [->, shorten >=1mm, shorten <=1mm] (PT1) to [out=60, in=120, looseness=8] (PT1);
        \end{scope}

        \node [draw, fill, below=3cm of P1] (T1) {}; \node [above=0cm of T1, font=\scriptsize] { 1 };
        \node [draw, fill, right=1cm of T1] (T3) {}; \node [above=0cm of T3, font=\scriptsize] { 3 };
        \node [draw, fill, below=0.7cm of T3] (T2) {}; \node [below=0cm of T2, font=\scriptsize] { 2 };
        \node [draw, fill, circle, right=1cm of T3, inner sep=2.5pt] (T4) {}; \node [above=0cm of T4, font=\scriptsize] { 4 };
        \node [draw, fill, diamond, below=0.7cm of T4, inner sep=2pt] (T5) {}; \node [below=0cm of T5, font=\scriptsize] { 5 };
        \draw [->, shorten >=1mm, shorten <=1mm, transform canvas={yshift=0.5mm}] (T1) -- (T3);
        \draw [<-, shorten >=1mm, shorten <=1mm, transform canvas={yshift=-0.5mm}] (T1) -- (T3);
        \draw [->, shorten >=1mm, shorten <=1mm] (T2) -- (T3);
        \draw [->, shorten >=1mm, shorten <=1mm] (T3) -- (T4);
        \draw [->, shorten >=1mm, shorten <=1mm] (T3) -- (T5);
        \node [above=0.7cm of T1.west, anchor=west, xshift=-0.2cm, font=\scriptsize\itshape] { tgt: };

        \begin{scope}[color=uofgcobalt]
        \node [draw, fill, below=2cm of T3] (TT1) {};
        \node [draw, fill, circle, inner sep=2.5pt, right=0.6cm of TT1] (TT2) {};
        \node [draw, fill, diamond, inner sep=2pt, left=0.6cm of TT1] (TT3) {};
        \draw [->, shorten >=1mm, shorten <=1mm] (TT1) -- (TT2);
        \draw [<->, shorten >=1mm, shorten <=1mm] (TT1) -- (TT3);
        \draw [->, shorten >=1mm, shorten <=1mm] (TT1) to [out=60, in=120, looseness=8] (TT1);
        \end{scope}
        \end{tikzpicture}\end{minipage}\begin{minipage}[c]{0.1\paperwidth}~\end{minipage}\begin{minipage}[c]{0.6\paperwidth}
        \only<2>{\lstinputlisting[basicstyle=\scriptsize\ttfamily]{retype-model.essence}}%
        \only<3>{\lstinputlisting[basicstyle=\scriptsize\ttfamily]{retype-constraints.essence}}%
        \only<4>{\lstinputlisting[basicstyle=\scriptsize\ttfamily]{retype-instance.essence}}%
        \only<5>{\lstinputlisting[basicstyle=\scriptsize\ttfamily]{retype-solutions.essence}}%
    \end{minipage}
\end{frame}

\begin{frame}{Costed Subgraph Isomorphism}
    \begin{itemize}
        \item We can look for the ``least expensive'' assignment:
    \lstinputlisting[basicstyle=\scriptsize\ttfamily]{costing.essence}
        \item Or something fancier like a Pareto front, or involving combinations of assignments.
    \end{itemize}
\end{frame}

\begin{frame}{Unfortunately\ldots}
    \only<1>{\centering\includegraphics{gen-graph-glasgow-versus-minion-cumulative.pdf}}%
    \only<2>{\centering\includegraphics{gen-graph-glasgow-versus-minion-scatter.pdf}}
\end{frame}

\begin{frame}{Communicating Solvers}
    \begin{itemize}
        \item The Glasgow Subgraph Solver is basically a constraint programming solver for graph problems.
            \begin{itemize}
                \item Special data structures and propagation algorithms.
                \item Would be a lot of work to add all the usual general purpose constraints.
                \item Not particularly easy to add arbitrary other variables.
            \end{itemize}
        \item Idea: use a general CP solver to handle side constraints and additional variables.
    \end{itemize}
\end{frame}

\begin{frame}{How to Communicate?}
    \begin{itemize}
        \item External CP solver has a model of everything except (possibly) the graph constraints.
        \item Glasgow solver feeds partial assignments and can ask:
            \begin{itemize}
                \item Propagate, and either tell me ``unsatisfiable'', or list any deletions inferred.
                \item Give me one or all solutions respecting this assignment.
            \end{itemize}
        \item Simple text protocol, communicate over named pipes.
            \begin{itemize}
                \item Implemented in Minion, but nothing particularly solver-specific.
            \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{High Level Modelling and Coordination}
    \begin{itemize}
        \item Main challenges are:
            \begin{itemize}
                \item We need the two solvers to have a consistent view of the graph variables.
                \item Recognising what the graph part of the problem is.
            \end{itemize}
        \item Essence toolchain is good at pattern matching.
        \item Minimal text file explaining variable remapping to the Glasgow solver.
    \end{itemize}
\end{frame}

\begin{frame}{When to Communicate?}
    \begin{itemize}
        \item Just check solutions?
        \item Propagate at the root node as well?
        \item Propagate at every node?
        \item Something cleverer?
    \end{itemize}
\end{frame}

\begin{frame}{Four Challenging Problems}
    \only<1>{\centering\includegraphics{gen-graph-nosideconstraints.pdf}}%
    \only<2>{\centering\includegraphics{gen-graph-oddeven.pdf}}%
    \only<3>{\centering\includegraphics{gen-graph-mostlyodd.pdf}}%
    \only<4>{\centering\includegraphics{gen-graph-parity.pdf}}%
\end{frame}

\begin{frame}{Rollback Communication}
    \begin{itemize}
        \item Propagate at the root node.
        \item Check solutions. If a solution is rejected:
            \begin{itemize}
                \item Backtrack and propagate.
                \item If this is rejected, backtrack and propagate again until success.
                \item Otherwise, continue without propagating.
            \end{itemize}
        \item Heavily inspired by backjumping, conflict analysis, and SMT, but doesn't require the CP
            solver to be able to produce conflict clauses.
    \end{itemize}
\end{frame}

\begin{frame}{Perspectives}
    \begin{itemize}
        \item For hard instances, this is great.
        \item Also extremely useful for prototyping and rapid data exploration.
        \item Very large startup overheads make this challenging for some other applications.
            \begin{itemize}
                \item But this is down to high level modelling rather than using a CP solver.
                \item Also questions about deployability\ldots
            \end{itemize}
        \item As is often the case, knowing when to do strong reasoning is critical.
    \end{itemize}
\end{frame}

{
    \usebackgroundtemplate{
        \tikz[overlay, remember picture]
        \node[at=(current page.south), anchor=south, inner
        sep=0pt]{\includegraphics[keepaspectratio=true, width=\paperwidth]{background2.jpg}};
    }

    \begin{frame}[plain,noframenumbering]
        \begin{tikzpicture}[remember picture, overlay]
            \node at (current page.north west) {
                \begin{tikzpicture}[remember picture, overlay]
                    \fill [fill=uofguniversityblue, anchor=north west] (0, 0) rectangle (\paperwidth, -2.6cm);
                \end{tikzpicture}
            };

        \node (logo) [anchor=north east, shift={(-2.2cm,-0.4cm)}] at (current page.north east) {
            \includegraphics[keepaspectratio=true,scale=0.4]{UoG_keyline.pdf}
        };
        \node (logo2) [anchor=north, below = 0 of logo.south] {
            \includegraphics[keepaspectratio=true,scale=0.025]{UStA_Logo_PR-REV.png}
        };
        \node (logo3) [anchor=west, right = 0 of $(logo2.east)!0.5!(logo.east)$] {
            \includegraphics[keepaspectratio=true,scale=0.1]{kings.png}
        };

            \node [anchor=north west, shift={(0.6cm,-0.8cm)}] at (current page.north west) {
                \textcolor{white}{\url{https://ciaranm.github.io/}}
            };

            \node [anchor=north west, shift={(0.6cm,-1.4cm)}] at (current page.north west) {
                    \textcolor{white}{\href{mailto:ciaran.mccreesh@glasgow.ac.uk}{\nolinkurl{ciaran.mccreesh@glasgow.ac.uk}}}
            };
        \end{tikzpicture}
    \end{frame}
}

\end{document}
